---
title: "Getting started with mosaik"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mosaik}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 5)
```

mosaik is a small set of operations on one kind of object: `syn_*` synthesises
fields, `mdf_*` modifies layers, `msr_*` measures them, and `msr()` combines
measurements into metrics. This vignette walks through them on the example data
that ships with the package.

```{r}
library(mosaik)
```

## The example data

`landscape` is a small example: a 60 by 56 grid with two layers, a land-cover
map (`cover`) and a continuous `intensity` field.

```{r}
landscape
```

```{r, fig.alt = "The cover layer of the example data, nine land-cover classes"}
msk_vis(landscape, .layer("cover"))
```

The cell values of `cover` are the class identifiers themselves: 1, 11, 21 and
so on. There is no lookup from a number to a class; the number is the class.

## Measuring the primitives

A `msr_*` function measures one primitive and writes the result into the
object at the level it describes. Here the area of every class, and of the
whole grid:

```{r}
m <- landscape |>
  msr_area(scale = "class", layer = "cover") |>
  msr_area(scale = "landscape", layer = "cover") |>
  msr_perimeter(scale = "class", layer = "cover")
```

Class-level results are in the categories of the layer, landscape-level results
in the global slot:

```{r}
msk_categories(m)$cover
msk_global(m)
```

The primitives are area, number, perimeter, adjacency, dissimilarity and cost,
each measured at patch, class or landscape level.

## Composing a metric

`msr()` computes a metric from an equation over primitives that have already
been measured. A primitive is named as `metric.scale`, so `area.class` is the
area of each class and `area.landscape` the area of the whole grid. The
share of the grid each class covers, and its edge density:

```{r}
m <- m |>
  msr(equation = "area.class / area.landscape * 100", label = "pland",
      layer = "cover") |>
  msr(equation = "perimeter.class / area.landscape", label = "edge_density",
      layer = "cover")

msk_categories(m)$cover
```

The result is stored where its length says it belongs: one value per class goes
to the class level, a single value to the landscape level. Shannon diversity
reduces over the classes, so it lands in the global slot:

```{r}
m <- msr(m, layer = "cover", label = "shannon",
         equation = "-sum(area.class / area.landscape *
                          log(area.class / area.landscape))")
msk_global(m)$shannon
```

No function was written for any of these metrics. That is the point of the
package: a metric that no catalogue ships is one more equation.

## Patches

Patch-level primitives work the same way. Here the forest class (47) is turned
into a layer of its own, and every forest patch is measured:

```{r}
f <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  msr_area(scale = "patch", layer = "forest") |>
  msr_perimeter(scale = "patch", layer = "forest") |>
  msr(equation = "perimeter.patch / sqrt(area.patch)", label = "shape_index",
      layer = "forest")
```

The patches are found once and kept in the object, so every patch-level
measurement refers to the same patches. After binarising, the non-forest
background is class 0, so the forest patches are the rows of class 1:

```{r}
p <- msk_patches(f)
patches <- data.frame(p[-3])
patches[patches$class == 1, ]
```

## Modifying layers

The `mdf_*` functions are generic operations such as binarise, dilate, erode,
mask and rescale. They do not know what a layer represents: eroding the forest
by one cell gives its core, the forest that is not edge, because erosion was
applied to forest, not because the operation knows about forests.

```{r, fig.alt = "Forest and its core after eroding by one cell"}
f <- mdf_erode(f, layer = "forest", add = "core")
msk_vis(f, .layer("forest"), .layer("core"))
```

## Every step is recorded

Each operation appends an entry to the object's provenance: what was done, with
which arguments, when, and by which version of mosaik.

```{r}
unlist(lapply(msk_provenance(f), names))
```

This record travels with the object, so a number read off it can be traced
back through every step that produced it.

## Your own data

A `SpatRaster` from terra becomes a mosaik object with `mosaik(rast = ...)`,
and `msk_terra()` turns one back into a `SpatRaster`. Everything above then
works unchanged.

```{r, eval = FALSE}
lc <- mosaik(rast = terra::rast("land_cover.tif"))
```

## Where next

- `vignette("measurement")`: why landscape condition has to be measured at the
  landscape level, and how mosaik approaches reference states and disclosure.
- `vignette("recipes")`: pipelines defined once, without data, and replayed on
  many rasters.
- `vignette("mspa")`: Morphological Spatial Pattern Analysis built from the
  primitives, matching the published classification cell for cell.
