---
title: "Getting started with mosaik"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mosaik}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 5)
```

mosaik follows the Unix principle for spatial analysis: each operator does
only one thing, but that as well as possible. Each function takes and returns
a mosaik, and an analysis is a combination of operators rather than a
function of its own. New questions need new
combinations, not new code, and every combination can be checked part by part
and rerun from its record.

The operators draw fields (`drw_*`), modify layers (`mdf_*`, `mdf()`) and
measure them (`msr_*`, `msr()`). This vignette takes each step in turn, on a
drawn field and on the example data that ships with the package.

```{r}
library(mosaik)
```

## Drawing a field

A mosaik starts as an empty grid. A `drw_*` function fills it with a field of
known structure: noise, a gradient, a regular pattern, clusters or a
tessellation. Here a random cluster field with four classes:

```{r, fig.alt = "A random cluster field with four classes"}
g <- mosaik(extent = c(0, 60, 0, 60), res = 1) |>
  drw_cluster(type = "randomCluster", p = 0.6, n = 4, seed = 1)
msk_vis(g)
```

Because the structure of such a field is known, it is a test case: a metric
can be checked on it before it is trusted on a real map.

## The example data

`landscape` is a small example: a 60 by 56 grid with two layers, a land-cover
map (`cover`) and the height of the vegetation on each cell (`canopy`, in metres).

```{r}
landscape
```

```{r, fig.width = 9, fig.alt = "The two layers of the example data: ten land-cover classes, and the canopy height in metres"}
msk_vis(landscape, .layer("cover"), .layer("canopy"))
```

The cell values of `cover` are the class identifiers themselves: 1, 11, 21 and
so on. There is no lookup from a number to a class; the number is the class,
and its label (47 is forest) is kept alongside, in the categories of the layer.

## Modifying layers

The `mdf_*` functions are generic operations such as binarise, dilate, erode,
mask and rescale. Each reads one layer and writes its result as a new layer
(`add`) next to it. Binarising the cover map at the forest class (47) gives a
layer that is 1 where there is forest and 0 elsewhere; eroding that by one cell
gives the forest core, the forest that is not edge.

```{r, fig.alt = "Forest and its core after eroding by one cell", fig.height = 3}
f <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_erode(layer = "forest", add = "core")
msk_vis(f, .layer("forest"), .layer("core"))
```

The operations do not know what a layer represents. Erosion gives the core
because it was applied to forest, not because it knows about forests. A
sequence of such steps can also be written down once, without data, and
replayed on any map with `mdf()`; `vignette("techniques")` shows how.

## Measuring the primitives

A `msr_*` function measures one primitive for every class of a layer and
writes the result into the class table of that layer. Here the area and
perimeter of every class of the cover map:

```{r}
m <- landscape |>
  msr_area(layer = "cover") |>
  msr_perimeter(layer = "cover")
```

Every result is stored with the layer it was measured on. `msk_categories()`
returns the class table of a layer:

```{r}
msk_categories(m, layer = "cover")
```

The same functions measure single patches, once the patches are numbered.
Which cells form a patch depends on whether cells that touch only at a corner
count as connected, so that is a step of its own: `mdf_componentise()`, with
four or eight neighbours. It writes a layer in which every cell carries the
number of its patch. Each class of that layer is a patch, so measuring the
layer measures the patches. On the forest layer, the patches are those of the
forest (1); the non-forest land (0) forms none:

```{r}
f <- f |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  msr_area(layer = "patch") |>
  msr_perimeter(layer = "patch")

p <- msk_categories(f, layer = "patch")
data.frame(patch = p$gid, area = p$area, perimeter = p$perimeter)
```

The primitives are area, perimeter, adjacency, distance and dissimilarity.

## Composing a metric

`msr()` computes a metric from an equation over values already in the object.
The equation is evaluated with one class in focus at a time, and a value is
named as `name.focus`: `area.self` is the area of the class in focus,
`area.others` the areas of all other classes, `area.all` the areas of all
classes. The share of the grid each class covers, and its edge density:

```{r}
m <- m |>
  msr(equation = "area.self / sum(area.all) * 100", label = "pland",
      layer = "cover") |>
  msr(equation = "perimeter.self / sum(area.all)", label = "density",
      layer = "cover")

data.frame(msk_categories(m, layer = "cover")[c("val", "pland", "density")])
```

With `.self` or `.others`, the equation is evaluated once for each class and
the results go to the class table; with only `.all`, it is evaluated once for
the whole layer. An equation that mixes them is evaluated once for each class:
`pland` above has `area.self`, so it runs once per class, and `sum(area.all)`
is the same total in every run. Each run gives the value of one class, or of
the whole layer, so it must come to one number. `area.self` is one
number, but `area.others` and `area.all` hold one number per class, so a
function such as `sum()` or `max()` has to reduce them. An equation without
that reduction, such as `area.all * 2`, has no single result, and `msr()`
reports it as an error. Shannon diversity reduces the shares of all classes to
one value:

```{r}
m <- msr(m, layer = "cover", label = "shannon",
         equation = "-sum(area.all / sum(area.all) *
                          log(area.all / sum(area.all)))")
msk_global(m, layer = "cover")$shannon
```

A metric computed this way can itself be used in the next equation;
`vignette("techniques")` shows how.

Patch-level metrics are the same equations on the layer of patches. The shape
index of every forest patch:

```{r}
f <- msr(f, equation = "perimeter.self / sqrt(area.self)", label = "shape",
         layer = "patch")
```

No function was written for any of these metrics. A metric that no catalogue
ships is one more equation.

## Every step is recorded

Each operation appends an entry to the object's provenance: what was done, with
which arguments, when, and by which version of mosaik. Printing the object
lists the steps at the end, each with the layer it read, the layer it wrote,
and the arguments that differ from the defaults:

```{r}
f
```

This record travels with the object, so a number read off it can be traced
back through every step that produced it.

## Your own data

A `SpatRaster` from terra becomes a mosaik object with `mosaik(rast = ...)`,
and `msk_terra()` turns one back into a `SpatRaster`. Everything above then
works unchanged.

```{r, eval = FALSE}
lc <- mosaik(rast = terra::rast("land_cover.tif"))
```
