---
title: "Getting started with mosaik"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mosaik}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 5)
```

mosaik follows the Unix principle for spatial analysis: each operator does one
thing, all of them take and return a mosaik, and an analysis is a combination
of them rather than a function of its own. New questions need new
combinations, not new code, and every combination can be checked part by part
and rerun from its record.

The operators draw fields (`drw_*`), modify layers (`mdf_*`, `mdf()`) and
measure them (`msr_*`, `msr()`). This vignette takes each step in turn, on a
drawn field and on the example data that ships with the package.

```{r}
library(mosaik)
```

## Drawing a field

A mosaik starts as an empty grid. A `drw_*` function fills it with a field of
known structure: noise, a gradient, a regular pattern, clusters or a
tessellation. Here a random cluster field with four classes:

```{r, fig.alt = "A random cluster field with four classes"}
g <- mosaik(extent = c(0, 60, 0, 60), res = 1) |>
  drw_cluster(type = "randomCluster", p = 0.6, n = 4, seed = 1)
msk_vis(g)
```

Because the structure of such a field is known, it is a test case: a metric
can be checked on it before it is trusted on a real map.

## The example data

`landscape` is a small example: a 60 by 56 grid with two layers, a land-cover
map (`cover`) and the height of the vegetation on each cell (`canopy`, in metres).

```{r}
landscape
```

```{r, fig.width = 9, fig.alt = "The two layers of the example data: ten land-cover classes, and the canopy height in metres"}
msk_vis(landscape, .layer("cover"), .layer("canopy"))
```

The cell values of `cover` are the class identifiers themselves: 1, 11, 21 and
so on. There is no lookup from a number to a class; the number is the class,
and its label (47 is forest) is kept alongside, in the categories of the layer.

## Modifying layers

The `mdf_*` functions are generic operations such as binarise, dilate, erode,
mask and rescale. Each reads one layer and writes its result as a new layer
(`add`) next to it. Binarising the cover map at the forest class (47) gives a
layer that is 1 where there is forest and 0 elsewhere; eroding that by one cell
gives the forest core, the forest that is not edge.

```{r, fig.alt = "Forest and its core after eroding by one cell", fig.height = 3}
f <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_erode(layer = "forest", add = "core")
msk_vis(f, .layer("forest"), .layer("core"))
```

The operations do not know what a layer represents. Erosion gives the core
because it was applied to forest, not because it knows about forests. A
sequence of such steps can also be written down once, without data, and
replayed on any map with `mdf()`; `vignette("techniques")` shows how.

## Measuring the primitives

A `msr_*` function measures one primitive and writes the result into the
object at the level it describes. Here the area and perimeter of every class
of the cover map, and the area of the whole grid:

```{r}
m <- landscape |>
  msr_area(scale = "class", layer = "cover") |>
  msr_area(scale = "landscape", layer = "cover") |>
  msr_perimeter(scale = "class", layer = "cover")
```

Every result is stored with the layer it was measured on. `msk_categories()`
returns the class-level results of a layer, `msk_global()` its landscape-level
results:

```{r}
msk_categories(m, layer = "cover")
msk_global(m, layer = "cover")
```

The same functions measure single patches, once the patches are numbered.
Which cells form a patch depends on whether cells that touch only at a corner
count as connected, so that is a step of its own: `mdf_componentise()`, with
four or eight neighbours. It keeps the patches in the object, so every
patch-level measurement refers to the same patches. On the forest layer, the
patches are those of the forest (1); the non-forest land (0) forms none:

```{r}
f <- f |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  msr_area(scale = "patch", layer = "forest") |>
  msr_perimeter(scale = "patch", layer = "forest")

p <- msk_patches(f, layer = "forest")
data.frame(patch = p$patch, area = p$area, perimeter = p$perimeter)
```

The primitives are area, number, perimeter, adjacency, dissimilarity and cost.

## Composing a metric

`msr()` computes a metric from an equation over values already in the object.
A value is named as `metric.scale`, so `area.class` is the area of each class
and `area.landscape` the area of the whole grid. The share of the grid each
class covers, and its edge density:

```{r}
m <- m |>
  msr(equation = "area.class / area.landscape * 100", label = "pland.class",
      layer = "cover") |>
  msr(equation = "perimeter.class / area.landscape", label = "edge_density.class",
      layer = "cover")

data.frame(msk_categories(m, layer = "cover")[c("val", "pland", "edge_density")])
```

The label is written like a variable: `pland.class` stores the result with the
classes, under the name `pland`, and `msr()` stops if the equation does not give
one value per class.

Shannon diversity sums over the classes into one value for the landscape:

```{r}
m <- msr(m, layer = "cover", label = "shannon.landscape",
         equation = "-sum(area.class / area.landscape *
                          log(area.class / area.landscape))")
msk_global(m, layer = "cover")$shannon
```

A metric computed this way can itself be used in the next equation;
`vignette("techniques")` shows how.

Patch-level primitives compose the same way. The shape index of every forest
patch:

```{r}
f <- msr(f, equation = "perimeter.patch / sqrt(area.patch)",
         label = "shape_index.patch", layer = "forest")
```

No function was written for any of these metrics. A metric that no catalogue
ships is one more equation.

## Every step is recorded

Each operation appends an entry to the object's provenance: what was done, with
which arguments, when, and by which version of mosaik. Printing the object
lists the steps at the end, each with the layer it read, the layer it wrote,
and the arguments that differ from the defaults:

```{r}
f
```

This record travels with the object, so a number read off it can be traced
back through every step that produced it.

## Your own data

A `SpatRaster` from terra becomes a mosaik object with `mosaik(rast = ...)`,
and `msk_terra()` turns one back into a `SpatRaster`. Everything above then
works unchanged.

```{r, eval = FALSE}
lc <- mosaik(rast = terra::rast("land_cover.tif"))
```
