---
title: "mosaik: one engine for spatially explicit landscape measurement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mosaik: one engine for spatially explicit landscape measurement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Abstract

Landscape measurement looks like a sprawl of incompatible methods — neutral
landscape models, FRAGSTATS indices, morphological pattern analysis,
connectivity models, ecosystem-service maps, land-change scenarios, condition
accounts. It is not. Almost every named metric in common use decomposes into a
small set of measurement primitives — area, number, perimeter, adjacency,
distance, dissimilarity — composed over a stack of raster layers. mosaik is
built on that decomposition rather than on a catalogue of named metrics: it
makes layers, modifies them, measures the primitives — directly or by combining
them into any metric — and simulates configurations toward targets. Because every method reduces to the
same primitives plus composition, one engine covers null-model generation,
metric computation, scenario building, and condition accounting — not by
bolting tools together, but because they were the same operations all along.

This unification is what opens computational landscape ecology to a wider
audience. The neutral-model builder, the landscape-metrics user, the
scenario modeller, and the analyst who must produce a defensible
biodiversity-condition number for disclosure today reach for four separate
tools with four vocabularies. mosaik gives them one open, inspectable,
provenance-tracked engine, and stays deliberately agnostic about the value
choices the numbers encode. That neutrality is the contribution: not a missing
metric, but a missing common substrate the field can build standards on before
those standards are locked in by proprietary practice.

This document explains what mosaik is, the problem it answers, how it is built,
and why it is built that way. For the function-level reference see the package
documentation; for whether mosaik fits your particular work, the closing
section points to the use-case vignettes.

## Introduction: the measurement gap

Biodiversity disclosure frameworks — the European Sustainability Reporting
Standards E4 (ESRS E4), the Taskforce on Nature-related Financial Disclosures
(TNFD), the Science Based Targets Network (SBTN) — specify *what* must be
reported but provide no rigorous basis for *how* landscape condition is
measured. The IPBES Business and Biodiversity Assessment (2026) records the gap
directly: current metrics are insufficient to track the changes in nature that
matter for business and society, and functional diversity beyond species
richness lacks adequate indicators.

In practice, corporate biodiversity reporting defaults to species counts at
points, land-cover proportions inside arbitrary buffers, and qualitative site
descriptions. These are measurable but ecologically thin: they treat
biodiversity as a property of discrete sites rather than as an outcome of
landscape-scale process. A forest patch surrounded by intensive agriculture is
functionally different from the same patch in continuous woodland, even when
internal species counts are identical.

What is absent is not data and not another metric. It is an open, interoperable
tool chain that runs from raw spatial data through landscape-structural
measurement to standards-aligned outputs with full provenance — and that stays
agnostic about the value choices the numbers encode. Existing global metrics
tend to bake one such choice into the metric itself. The Ecosystem Integrity
Index, for example, measures functional condition against a potential-natural
baseline; that is a defensible choice, but it is *a* choice, written into the
metric's definition rather than left open to the user. mosaik takes the
opposite stance: it supplies the engine and lets the reference, the objective,
and the vocabulary be set by the user or the standard, not by the tool. That is
the gap mosaik exists to close.

## One engine: primitives and four verbs

The decomposition is the whole design. Shannon diversity, contagion, edge
density, mean patch size, evenness, a morphological class: each is one of a few
primitives, or an arithmetic combination of them, computed over a stack of
raster layers. mosaik measures the primitives and lets the user build any
named metric from them. Four families of verbs follow:

| Verb | Family | Role |
| :--- | :--- | :--------- |
| **Make** | `make_*` | materialise a named layer — pattern primitives and layers computed from inputs — or ingest real ones |
| **Modify** | `mdf_*`, `mdf()` | transform or combine existing layers without claiming new semantics; `mdf()` replays a recorded sequence of such steps (a *recipe*) so an algorithm defined once applies to many rasters |
| **Measure** | `msr_*`, `msr()` | `msr_*` measures the primitives directly (area, number, perimeter, adjacency, distance, dissimilarity) at patch, class, and landscape level; `msr()` measures a derived metric by combining those primitives via an equation, written into the matching slot |
| **Simulate** | `sim` | evolve a configuration over time toward targets defined on any `msr_*` output |

Every family takes a `mosaik` object in and returns one out, so operations
compose into pipelines with provenance carried through each step. The data
structure is one S4 class with transparent, inspectable slots — `extent`,
`dims`, `layers`, `categories`, `patches`, `global`, `crs`, `provenance` — and
no opaque pointers. Cell values *are* the group identifiers; metric outputs land
in the slot matching their level (patch-level in `@patches`, class-level in
`@categories`, landscape-level in `@global`). Two consequences fall out at once:

- Morphological Spatial Pattern Analysis (core, edge, bridge, loop, branch,
  islet) is a measurement over the configuration primitive, not a separate
  program to wrap. It belongs in `msr_*`.
- The FRAGSTATS family of landscape indices is not special. Each is a named
  combination of the primitives, computed by the same machinery — which is why
  users can construct metrics that no fixed catalogue ships.

Work begins with a grid. `mosaik()` creates an empty one (or ingests a real
raster); `syn_*` functions populate it with abstract fields — noise, texture,
gradients, tessellations — all of which take a mosaik in and return one out, so
they chain like everything else:

```{r}
library(mosaik)

# a synthetic field to analyse
m <- mosaik(extent = c(0, 100, 0, 100), res = 1) |>
  syn_texture(type = "diamondSquare", name = "elevation", role = "surface")
```

Whole synthetic *landscapes* — climate, soil, vegetation, land use, tenure, each
derived from the last — are the business of the companion package
[mundus](https://github.com/resonantia-earth/mundus), which builds on this class
and returns the same objects. Point the engine at a real raster
(`mosaik(rast = ...)`) instead and nothing downstream changes: measured and
generated layers sit side by side and every later operation treats them
identically.

A minimal status-quo assessment on a real landscape makes the measurement shape
concrete:

```{r}
mosaik(rast = land_cover) |>
  msr_area(scale = "class") |>
  msr_perimeter(scale = "class") |>
  msr_area(scale = "landscape") |>
  msr(equation = "perimeter.class / area.class",      label = "edge_density") |>
  msr(equation = "area.class / area.landscape * 100", label = "prop_area")
```

This is the difference from a metrics catalogue such as landscapemetrics, which ships a fixed set of pre-baked indices: you can compute the ones it includes, but not ones it does not. By exposing the primitives and letting `msr()` combine them, mosaik lets you build any named metric — including ones no catalogue ships.

Because every step records itself in the object's provenance, a sequence of
steps can also be defined *without* a starting raster and replayed later on
many. Calling the operators with no input mosaik records the chain rather than
executing it; the result is a **recipe**, and `mdf()` applies it to a real
mosaik:

```{r}
# define a morphological algorithm once, with no data
gaps <- mdf_componentise(layer = "fg", add = "cc_fore") |>
  mdf_componentise(layer = "bg", add = "cc_back") |>
  mdf_dilate(layer = "fg", add = "closed") |>
  mdf_mask(layer = "closed", by = "cc_back", add = "gaps")

# apply it to any landscape
result <- mdf(some_landscape, gaps)
```

A step reads an earlier intermediate by the layer name an earlier step wrote
(`by = "cc_back"`), so branch-and-recombine algorithms — gap detection, core
habitat, corridor identification — are expressed as named layers rather than
hand-written loops, and the recipe carries its own provenance for audit. This
is the analyst's counterpart to a fixed metric catalogue: not a menu of
canned procedures, but a way to capture and reuse your own.

A fourth verb, `sim`, runs the same object forward in time. It lives in
[mundus](https://github.com/resonantia-earth/mundus) rather than here, but it is
worth seeing because it consumes exactly what this package measures: given land
types and demand targets, it reallocates cells over a number of steps until the
targets are met — a scenario explorer driven by the same `msr_*` outputs the
analysis side computes:

```{r}
types <- sim_landtype(name = "forest", produces = list(timber = 5, carbon = 10),
                      allow = list(crop = list(strategy = sim_strategy("frontier")))) |>
         sim_landtype(name = "crop", produces = list(food = 8))
targets <- sim_target(name = "food",   target = 1000, match = "minimum") |>
           sim_target(name = "carbon", target = 5000, match = "minimum")

scenario <- sim(land_cover, types = types, targets = targets, n = 10)
```

Run it once for a single scenario, or many times with `noise > 0` to obtain a
*distribution* of reachable landscapes — the reference ensemble the next section
builds on.

## Why mosaik has the shape it does

Five arguments motivate the design. Each is load-bearing for a different reader,
with the core arguments below and detailed discussion elsewhere (linked).

### Why measurement must be at the landscape level

Three independent reasons, none reducible to the others.

*Additionality and leakage are landscape properties.* A hectare of restored
forest may be additional (a genuine gain over what would otherwise have
happened) or non-additional (it would have regenerated anyway); the same hectare
may displace conversion to a neighbouring parcel and net the gain to zero. Both
require knowing the spatial alternatives the parcel could have produced and the
configuration of surrounding land. A point assessment cannot reach them; a
landscape assessment, with a defensible reference distribution, can.

*Ecological function is carried by structure.* A 100 ha forest can be one block,
ten blocks, or ten blocks with corridors — same area, radically different
biodiversity outcomes. Patch shape, edge length, inter-patch distance, and the
intervening matrix decide whether an area number translates into population
persistence. Polygon-level reporting flattens this; cell-level structural
metrics preserve it.

*Temporal claims need a landscape baseline.* "Nature positive" is
time-relative — positive compared to what? The candidate baselines all fail, for
reasons developed next.

### The reference-state reframing

Condition accounting needs a reference: "good condition" relative to what? Every
candidate fails. A historical baseline forces an unanswerable question — which
year? Potential natural vegetation is a construct that may never have existed as
a stable state. Least-disturbed reference sites are biased toward whatever
happens to be left. The reason none works is that the question "what should this
landscape look like" is normative, not empirical. No amount of better data
resolves a value choice.

mosaik's response is to stop hiding the value choice inside the word "natural"
and make it explicit — then let the user, not the tool, set the reference. One
powerful option this opens is to replace a fixed reference state with a
*distribution of reachable configurations*: given a landscape's constraints,
`sim` generates many configurations that satisfy them, and the distribution
of metric values across that ensemble is the reference. An observed landscape is
scored against the ensemble rather than against a single engineered number. This
sidesteps the baseline problem (no fixed reference need be named) and the
baseline-gaming that broke carbon markets (the reference is a distribution of
plausible landscapes, not a number a developer can quietly engineer toward).
mosaik does not impose this — it makes it constructible, alongside whatever
reference a standard or user prefers. It is the same epistemic move the
neutral-landscape-model community already makes, with a constrained optimiser in
place of a randomiser.

Letting each user set their own reference raises an obvious objection:
cross-company comparability collapses if everyone picks a different objective.
The resolution is not to deny that the choice is normative but to *make it so
explicitly*. A standards body can prescribe a fixed objective function — a set
of indicators with fixed weights — while acknowledging openly that this is a
value choice, not a scientific fact. Companies then disclose against the
prescribed objective; auditors check whether the landscape actually delivers it;
researchers run alternative objectives to inform the next round of
standard-setting. mosaik supports all three roles on one engine. This is a
position the existing standards do not yet take, and it is mosaik's contribution
to disclosure methodology: not a metric, but a defensible way to fix one.

### Simple rules over process models

Where a quantity must be modelled rather than measured, mosaik occupies a
deliberate middle ground between adjacent-cell rules and full mechanistic
process models: a *budget model*. Track where material enters (deposition maps,
intensity proxies), follow flow direction (via `make_rivers`), apply a per-cell
retention fraction from the empirical literature, and accumulate at endpoints.
Structure does the work; the rules stay simple.

This is defensible for disclosure (physically grounded, transparent,
reproducible), fast at landscape scale (no iterative solver), and honest about
what it omits (no seasonal dynamics, no explicit biological uptake, no
groundwater — stated plainly). Full process models pay a high parameter cost
because of system coupling, not modeller pedantry, and are wrong outside their
parameterised regime; adjacent-cell rules break the moment a process couples
non-locally (nitrogen routed 20 km down a stream network, colonisation depending
on the whole configuration between source and target). The budget model is the
honest level of detail for landscape-scale assessment. The guiding rule: do not
make rules do what structure should do.

### Pressure layers

Some inputs the disclosure use case needs have no generative process — the data
already exist. Pressure proxies are the clearest case: road density
(OpenStreetMap), nighttime light (VIIRS), agricultural intensity, the Human
Modification Index. These enter mosaik as ordinary user-supplied layers, where
they can be combined into composite pressure scores or used to steer a
simulation. No pressure generator is needed. The one discipline this demands:
they are *activity proxies*, not measured ecosystem stresses — acceptable for 
comparative exposure ranking, not for precise causal attribution, and they must 
be labelled as such.

### What biodiversity does — the question underneath

The deepest gap is not tooling but theory, and it is worth stating because it
bounds what any tool can honestly claim. Underneath every metric choice sits a
question the field systematically avoids: what is biodiversity *for*, in the
sense of which functional capacity we are trying to preserve — resilience to
disturbance, continued service delivery, evolutionary potential, trophic
structure? Each implies different indicators, scales, and dynamics. The dominant
corporate metric, Mean Species Abundance, is spatially implicit and insensitive
to landscape structure, yet persists because it is implemented and changing it
disrupts existing reporting infrastructure. Patch-level landscape metrics are a
necessary step beyond it but are themselves insufficient without process
linkage. What is missing is *indicator theory*: a principled framework for which
structural metrics are reliable proxies for which ecological processes under
which conditions.

mosaik does not solve this — no software can. What it provides is the
infrastructure to attack it empirically: generate synthetic landscapes with
controlled structure, measure them reproducibly, compare assessments across
sites and scenarios, and demonstrate concretely what rigorous condition
assessment looks like and how far current frameworks fall short. The gap between
what frameworks accept as sufficient and what landscape science could deliver is
itself a finding — and one that needs to be visible before biodiversity markets
scale.

## Where mosaik fits: the disclosure stack

Six systems carry biodiversity disclosure today; none covers the full chain
alone. They form a stack, and mosaik occupies one link the stack does not
currently produce.

| Layer | Standard | What it classifies |
|---|---|---|
| Activity | NACE / ISIC | economic sectors and production processes |
| Activity → nature bridge | ENCORE | which services each process depends on, which pressures it exerts |
| Ecosystem services | CICES | provisioning, regulation & maintenance, cultural contributions |
| Ecosystem extent | IUCN GET | realms → biomes → functional groups |
| Accounting glue | SEEA-EA | combines extent, services, and condition into ecosystem accounts |
| Disclosure | TNFD, SBTN, GRI, CSRD/ESRS E4 | consume the layers above and produce disclosures |

The stack stops at the polygon — a sourcing region, an administrative unit, a
property boundary. mosaik operates one level down, at the cell, and supplies the
spatially explicit evidence the polygon-level numbers are supposed to rest on:

```
ENCORE: "crop production depends on pollination"
  → CICES 2.2.2.1: "pollination"
  → mosaik: cell-level condition and structure, scored against a reachable ensemble
  → SEEA-EA: aggregated to the administrative unit
  → TNFD / CSRD: a line in the report
```

mosaik produces the substrate; an orchestrator layer above it — built
separately — aggregates the cells upward and shapes them into a framework's
format. The package stays vocabulary-neutral: a land type that `produces` named
goods accepts arbitrary strings, not a fixed enumeration, and ontology codes
(CICES, IUCN GET) live in the `@categories` slot as a convention the user
maintains, carried through provenance but never validated by mosaik.
Hard-coding any one standard would tie the package to a single disclosure regime
and break its use in research and teaching.

A single mechanism covers the service side of this chain. `make_service`, in
mundus,
materialises a per-cell service layer from category-attached rates and spreads
it with a distribution kernel chosen by argument: isotropic decay for local
services (pollination, microclimate), flow-routed for transported resources
(nutrients down a stream network), point-to-point for first-order extraction
(timber forest → sawmill), or a vertical sink that accumulates into a
non-spatial scalar. This collapses what would otherwise be many bespoke
per-service models (one per service, in the manner of InVEST) into one call.

This works because mosaik does not model environmental realms as separate
compartments — there is no "atmosphere" or "anthroposphere" object. Each
resource that crosses a realm boundary is simply another layer (a vertical sink
accumulates into a `@global` scalar rather than into a hidden domain), so the
realms are implicit in which layers exist and how they connect. Everything stays
an inspectable layer, consistent with mosaik's preference for transparent
structure over opaque domains.

## Scope

mosaik covers spatial production up to first-order extraction and stops there.
It resolves spatial production into cell-level outputs; it does not resolve a
product into its supply-chain inputs.

| Tier | Example | Coverage |
|---|---|---|
| 1. Local biophysical services | pollination, soil retention, microclimate | mosaik |
| 2. Transported biophysical resources | water along rivers, sediment, nutrients | mosaik |
| 3. First-order extraction | timber forest → sawmill, ore mine → smelter | mosaik |
| 4. Second-order industrial processing | iron + coal → steel | out of scope — life-cycle assessment |
| 5. Aspatial service goods | software, finance | out of scope — input–output footprinting |

Beyond Tier 3 the right tools are life-cycle and input–output systems (openLCA,
ecoinvent, EXIOBASE, GLORIA), which meet mosaik at the orchestrator layer.

What mosaik is *not*, stated so the boundary can be enforced: it does not render
reports, validate classification vocabularies (CICES, EUNIS), maintain an
activity–service mapping (ENCORE), or shape outputs into a TNFD/CSRD/SBTN
report — those belong to the orchestrator above it. It is not a life-cycle or
input–output tool. And it is not a physical predictive simulator: `sim`
evolves a configuration toward targets under explicit strategies — a
scenario explorer, not a physics engine — and no conservation laws are
enforced unless the user wires them in. Absorbing any of these
would make mosaik either redundant (a worse copy of a better-resourced tool) or
unauditable (mixing scopes that need separate governance).

## Discussion: what mosaik solves, and what it leaves open

Rigorous landscape-condition measurement remains unsolved for four reasons;
mosaik addresses one.

1. **Normative.** Reference states and trade-off weights are value choices, not
   empirical facts. Better methods make them explicit and auditable — the
   reframing above. mosaik's contribution.
2. **Institutional.** Statistics offices, conservation agencies, and financial
   regulators built their frameworks independently; harmonisation is political,
   not technical.
3. **Data.** Condition accounts need in-situ validation at resolutions most
   monitoring programmes do not provide.
4. **Tooling.** No open, interoperable chain runs from raw spatial data to
   standards-aligned outputs with full provenance while staying agnostic about
   the value choices it encodes. **This agnostic substrate is the one mosaik
   supplies.** The other three are not mosaik's to solve.

mosaik does not settle which structural metrics are reliable proxies for which
ecological outcomes — the indicator-theory problem named above. It provides the
infrastructure to test that empirically, and demonstrates concretely what
rigorous assessment looks like. What is still owed, and where resonantia
engages, is the indicator theory itself and the work of engaging disclosure
frameworks before measurement conventions are locked in by commercial practice.

A further direction, not yet committed to, is that mosaik's primitives may not
be fundamentally spatial: area, number, adjacency, distance, and their
composition have natural interpretations for abundance matrices, time series,
and supply-chain networks, which would let the same measurement-and-composition
machinery serve non-spatial disclosure domains. That generalisation is explored
separately and is not a near-term commitment.

## Who this is for

mosaik has several front doors; each community arrives with a different question
and gets a use-case vignette that answers it without requiring this whole
document.

- **You build neutral landscape models** (NLMR, nlmpy). mosaik gives you
  process-based nulls — generated by a mechanism rather than a random field —
  and an optimiser that can hold structural targets fixed while varying the
  rest.
- **You compute landscape metrics or morphology** (FRAGSTATS,
  landscapemetrics, GUIDOS/MSPA). mosaik computes the same patch-, class-, and
  landscape-level metrics reproducibly inside a provenance-tracked pipeline, and
  treats morphology as one more measurement over the configuration primitive.
- **You build land-change scenarios** (the CLUE family, CLUMondo, Dinamica).
  Those models allocate area to satisfy a demand vector; `sim` allocates the
  same way and is being extended to optimise structural targets — patch-size
  distribution, connectivity — alongside area-to-meet-demand.
- **You work in nature finance or disclosure** (ESRS E4, TNFD, SBTN;
  biodiversity credits). You need a condition reference you can defend to an
  auditor and spatial computation you can run in the same place. mosaik lets you
  build the reference yourself rather than inherit one fixed inside a metric.

## A worked example of the claim

The argument above is that named methods decompose into a small set of
operators. The [MSPA vignette](mspa.html) carries that out in full for one of
them: morphological spatial pattern analysis, built from eight operators plus
`mdf_loop()`, with no function written for the method. Its output matches the
classification published with the original paper cell for cell, on the pattern
that paper introduced (shipped as the `mspa` dataset).

It is worth reading if you want to see what "compose it yourself" looks like in
practice, including where the composition is awkward.

*Maintained by resonantia research gUG (haftungsbeschränkt), Leipzig.*

## References

- Hobbs, R. J., Higgs, E., Harris, J. A. (2009). Novel ecosystems: implications
  for conservation and restoration. *Trends in Ecology & Evolution*, 24(11),
  599–605.
- Ellis, E. C. (2015). Ecology in an anthropogenic biosphere. *Ecological
  Monographs*, 85(3), 287–331.
- IPBES (2019). *Global Assessment Report on Biodiversity and Ecosystem
  Services*.
- IPBES (2026). *Business and Biodiversity Assessment* (IPBES-12, Manchester,
  February 2026).
- Kupfer, J. A. (2011). Theory in landscape ecology and its relevance to
  biogeography. In *SAGE Handbook of Biogeography*, ch. 4.
- Li, H., Wu, J. (2004). Use and misuse of landscape indices. *Landscape
  Ecology*, 19, 389–399.
- McGarigal, K., Marks, B. J. (1995). *FRAGSTATS: Spatial Pattern Analysis
  Program for Quantifying Landscape Structure*. USDA GTR PNW-GTR-351.
- UN (2021). *System of Environmental-Economic Accounting — Ecosystem
  Accounting* (SEEA-EA).
- Venter, O., et al. (2016). Global terrestrial Human Footprint maps for 1993
  and 2009. *Scientific Data*, 3, 160067.
