---
title: "Morphological spatial pattern analysis, built from operators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphological spatial pattern analysis, built from operators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 6)
```

## What this vignette shows

Morphological spatial pattern analysis (MSPA) divides the habitat of a binary
map into seven classes. Every habitat cell gets exactly one. The classes
describe the shape of the habitat, not its amount: which parts are interior,
which are the border towards the open land, and which connect one part with
another.

mosaik has no MSPA function, and it needs none. MSPA is a sequence of ordinary
operations: a distance to the nearest open cell, a threshold, numbering the
patches, a summary per patch and a thinning to centre lines. mosaik has all of
them. This vignette builds the classification from those parts and arrives at
the published result, cell for cell.

If you only want the classification, skip to the section with the whole
classification in one function and copy the code. The section after it runs
the method on an ordinary map and looks at what it makes of it.

## The seven classes

Every class is defined in relation to the **core**: habitat cells that lie
further than a chosen distance from the open land. That distance, `size`, is
the only real setting of the method. It says how deep inside the habitat a
cell must lie to be safe from what happens at the border.

The other six classes say how the rest of the habitat relates to the core:

| class | definition | reading |
|---|---|---|
| core | further than `size` from the open land | interior habitat |
| edge | the outer border of a core area | the border towards the open land |
| perforation | the border of a hole inside a core area | a border inside the habitat |
| islet | a patch without any core | too small for interior; a stepping stone |
| bridge | a strip joining **two different** core areas | a corridor |
| loop | a strip leaving a core area and returning to it | a detour, not a corridor |
| branch | a strip attached at one end only | a dead end |

Most of what MSPA says about connectivity lies in the difference between bridge
and loop. A bridge links two blocks of habitat, and losing it splits them. A
loop returns to where it started, and losing it splits nothing.

## Setup

The `mspa` dataset in the package is the pattern of the paper that introduced
the method (Soille & Vogt 2009), in its layer `pattern`. The paper prints the
classification of this pattern as its Fig. 1, so the map at the end of this
vignette can be held against it.

```{r setup, message = FALSE}
library(mosaik)

m <- msk_label(mspa, labels = data.frame(id = c(0, 1),
                                         label = c("open land", "habitat"),
                                         colour = c("#ffffff", "#222222")),
               layer = "pattern")
```

The habitat is the layer `pattern`, and every layer built below is added next
to it, so the object holds all intermediate results. This is the starting
point:

```{r input, echo = FALSE}
msk_vis(m, .layer("pattern"))
```

Two settings. `size` is the width of the border, the depth at which habitat
counts as interior. `edu` is the unit it is measured in. The reference
implementation fixes it at `sqrt(2)`, the distance between two cells that
touch at a corner, so `size = 1` means "further than `sqrt(2)` cell widths from
the open land".

```{r params}
size <- 1
edu  <- sqrt(2)

conn   <- 8L   # habitat cells touching at a corner are connected
connBg <- 4L   # open cells touching only at a corner are not
```

Each class is stored as a layer of its own, 1 where the class is and 0
elsewhere. `mdf_filter()` writes these layers: it takes a condition over the
layers and writes 1 where the condition holds and 0 where it does not.

Each class has a code, a name and a colour, taken from the paper so that the
maps here can be compared with its Fig. 1. Code 8 is not a class of the
method. It marks habitat that is not classified yet, drawn grey, so the maps
below show what is still to be explained. Code 9 marks the connecting strips
before they are sorted into bridge and loop.

```{r key}
classes <- c(branch = 7, edge = 6, perforation = 3, islet = 4,
             core = 1, bridge = 5, loop = 2)

key <- data.frame(
  id     = 0:9,
  label  = c("open land", "core", "loop", "perforation", "islet", "bridge",
             "edge", "branch", "not yet classified", "connecting strip"),
  colour = c("#ffffff", "#399c39", "#f7ff05", "#0401ed", "#9a3312",
             "#f10606", "#010100", "#efce09", "#cccccc", "#f10606"))
```

`show_classes()` draws a vector of class codes as a map on the grid of `obj`,
labelled with the names and colours of `key`. `show_state()` uses it to draw
the classes found so far and leaves the rest of the habitat grey. With every
step the grey shrinks.

```{r show-state}
show_classes <- function(obj, codes){
  g <- msk_add(msk_select(obj, pattern), mspa = as.numeric(codes))
  g <- msk_label(g, labels = key[key$id %in% codes, ], layer = "mspa")
  msk_vis(g, .layer("mspa"))
}

show_state <- function(obj, done = character()){
  codes <- ifelse(msk_pull(obj, "pattern") == 1, 8L, 0L)
  for(nm in done) codes[msk_pull(obj, nm) == 1] <- classes[[nm]]
  show_classes(obj, codes)
}
```

## Core

`mdf_distance()` gives every habitat cell its distance to the nearest open
cell. The core is what lies further than the threshold.

```{r core}
m <- mdf_distance(m, source = "background", layer = "pattern", add = "edt")
m <- mdf_filter(m, edt > size * edu, add = "core")
```

The green cells are core. What is still grey lies within `size` of the open
land, or belongs to a patch too narrow to hold any core. The rest of the
method tells these cases apart.

```{r core-fig, echo = FALSE}
show_state(m, "core")
```

## Islet

An islet is a patch without core. `mdf_componentise()` numbers the patches,
and for each patch the question is whether any of its cells is core.

`mdf_summarise()` answers that. It groups the cells by a layer of numbers and
writes one summary value onto every cell of the group, like a group-by in a
table. Here the summary is whether the patch holds a core cell, and the
patches without one are the islets.

```{r islet}
m <- mdf_componentise(m, connectivity = conn, layer = "pattern", add = "cc")
m <- mdf_filter(m, pattern != 0, value = TRUE, layer = "cc") |>
    mdf_replace(old = NA, new = 0, layer = "cc")
m <- mdf_summarise(m, by = "cc", fun = "any", layer = "core",
                   background = 0, add = "hasCore")
m <- mdf_filter(m, pattern == 1 & hasCore == 0, add = "islet")
```

The islets are settled: without core they have no border of a core area and
connect nothing.

```{r islet-fig, echo = FALSE}
show_state(m, c("core", "islet"))
```

## Edge and perforation

Both are the border of a core area, one cell wide. They differ in what they
face. An **edge** faces the open land around the habitat, a **perforation**
faces a hole inside a core area.

One test cannot tell them apart, because a patch can lie inside the hole of
another patch, and the border of that inner patch is an edge again. The
reference implementation therefore works from the outside in: it takes the
border of what is left, removes what lies inside a hole, and repeats, one
level of nesting per round.

The starting point: filling the holes of the core gives `corefill`, and what
the filling added are the holes themselves, `holes`.

```{r edges-init}
m <- mdf_distance(m, source = "foreground", layer = "core", add = "dcore")
m <- mdf_filter(m, core == 1 | (dcore > 0 & dcore <= size * edu), add = "opening")

m <- mdf_offset(m, value = 0, layer = "core", add = "work")
m <- mdf_fill(m, connectivity = connBg, layer = "work", add = "corefill")
m <- mdf_filter(m, corefill == 1 & work == 0, add = "holes")

# an empty layer to collect the edges in: no cell has pattern < 0
m <- mdf_filter(m, pattern < 0, add = "edge")
```

One round adds the border of the current `corefill` to `edge` and then removes
what lies inside holes, so that the next round sees the next level down.
A chain of `mdf_*` calls without a mosaik at its start does not run; it is
recorded as a **recipe**. `mdf_loop()` runs a recipe again and again until a
condition holds.

```{r edges-loop}
peel <- mdf_distance(source = "foreground", layer = "corefill", add = "dfill") |>
  mdf_filter(edge == 1 | (corefill == 0 & dfill > 0 & dfill <= size * edu), add = "edge") |>
  mdf_fill(connectivity = conn, layer = "holes", add = "i0") |>
  mdf_filter(corefill == 1 & i0 == 0, add = "corefill") |>
  mdf_filter(work == 1 & corefill == 0, add = "work") |>
  mdf_fill(connectivity = connBg, layer = "work", add = "corefill") |>
  mdf_filter(corefill == 1 & work == 0, add = "holes")
```

```{r edges-converge}
if(sum(msk_pull(m, "corefill")) > 0){
  m <- mdf_loop(m, peel, until = sum(corefill, na.rm = TRUE) == 0)
}
```

`mdf_loop()` checks its condition after each round. The `if` covers a map
without any core, on which no round should run at all.

The number of rounds depends on how deeply the pattern nests. On this pattern
the first round already collects all `r sum(msk_pull(m, "edge"))` border
cells, because nothing lies inside the hole of something else. A patch inside
a clearing inside a forest would need a second round.

Now the two are separated. What the rounds collected, minus the core itself,
is edge. The rest of the border of the core is perforation: the rounds never
reached it, because it faces inwards.

```{r edges-split}
m <- mdf_filter(m, edge == 1 & core == 0, add = "edge")
m <- mdf_filter(m, core == 0 & dcore > 0 & dcore <= size * edu & edge == 0,
                add = "perforation")
```

The core areas now have their full border. What is still grey touches no core
area on any side: it lies between them.

```{r edges-fig, echo = FALSE}
show_state(m, c("core", "islet", "edge", "perforation"))
```

## Connecting strips

The rest of the habitat is made of strips between the parts found so far.

```{r residues}
m <- mdf_filter(m, pattern == 1 & core == 0 & islet == 0 & perforation == 0 &
                  edge == 0, add = "residues")
```

To find which of these strips connect, the habitat is thinned to its centre
lines while the core areas stay as they are. That is
`mdf_skeletonise(anchor = "core", method = "homotopic")`. It removes a cell
only if removing it neither splits a piece of habitat in two nor opens a hole
in it.

That rule does the sorting. A strip that joins two core areas cannot be
removed without splitting the habitat, so its centre line remains. A strip
that ends nowhere can be removed entirely, so it disappears. No separate test
is needed to tell the two apart.

```{r skeleton}
m <- mdf_filter(m, opening == 1 | residues == 1, add = "skin")
m <- mdf_skeletonise(m, anchor = "core", background = 0, layer = "skin",
                     add = "sk", method = "homotopic")
m <- mdf_filter(m, sk == 1 & core == 0, add = "sk")
```

Of the centre lines, only those that can be reached through unclassified cells
are kept. They are found by growing a layer step by step until it no longer
changes, which is `mdf_loop(stable = TRUE)`.

```{r connector}
m <- mdf_filter(m, (opening == 1 & core == 0) | residues == 1, add = "i0")

m <- mdf_offset(m, value = 0, layer = "sk", add = "rec")
grow <- mdf_dilate(struct = msk_struct("square", width = 3), layer = "rec") |>
        mdf_filter(rec == 1 & i0 == 1, add = "rec")
m <- mdf_loop(m, grow, stable = TRUE, layer = "rec")

m <- mdf_filter(m, sk == 1 & i0 == 1 & rec == 1, add = "connector")
```

These are the connecting strips of the pattern, drawn over the grey they come
from. What remains is to say what each of them connects.

```{r connector-fig, echo = FALSE}
codes <- ifelse(msk_pull(m, "pattern") == 1, 8L, 0L)
codes[msk_pull(m, "connector") == 1] <- 9L
show_classes(m, codes)
```

## Bridge, loop and branch

Number each strip and each core area, then ask two questions of every strip:
how many different core areas it touches, and how many cells it has. Both are
summaries by `mdf_summarise()`; the first looks at the cells next to each strip
rather than at the strip itself.

A strip that touches two or more core areas is a **bridge**, if it is longer
than one cell. The second condition matters: a single cell can sit diagonally
in the gap between two core areas and touch both without running between
them, and the reference implementation counts it as a loop.

```{r bridge-loop}
m <- mdf_componentise(m, connectivity = conn, layer = "connector", add = "strand")
m <- mdf_filter(m, connector != 0, value = TRUE, layer = "strand") |>
    mdf_replace(old = NA, new = 0, layer = "strand")
m <- mdf_componentise(m, connectivity = conn, layer = "core", add = "coreID")
m <- mdf_filter(m, core != 0, value = TRUE, layer = "coreID") |>
    mdf_replace(old = NA, new = 0, layer = "coreID")

m <- mdf_summarise(m, by = "strand", fun = "n_distinct", neighbours = TRUE,
                   connectivity = conn, layer = "coreID", background = 0,
                   add = "nCores")
m <- mdf_summarise(m, by = "strand", fun = "n", layer = "strand",
                   background = 0, add = "strandSize")

m <- mdf_filter(m, connector == 1 & nCores >= 2 & strandSize >= 2, add = "bridge")
m <- mdf_filter(m, connector == 1 & bridge == 0, add = "loop")
```

This is the difference the method exists for: losing a bridge separates two
blocks of habitat, losing a loop does not.

```{r bridge-loop-fig, echo = FALSE}
show_state(m, c("core", "islet", "edge", "perforation", "bridge", "loop"))
```

Branch is what is left: the strips the thinning removed because they lead
nowhere.

```{r branch}
m <- mdf_filter(m, pattern == 1 & core == 0 & islet == 0 & perforation == 0 &
                  edge == 0 & bridge == 0 & loop == 0, add = "branch")
```

No grey is left. Every habitat cell now has a class.

```{r branch-fig, echo = FALSE}
show_state(m, names(classes))
```

## Assembling

The seven layers are combined into one map of class codes. Where two classes
claim the same cell, the order decides, because a cell can be both border and
connecting strip. The strips win, so bridge and loop are written last.

```{r assemble}
cls <- rep(0L, msk_ncells(m))
for(nm in names(classes)) cls[msk_pull(m, nm) == 1] <- classes[[nm]]
cls[msk_pull(m, "pattern") != 1] <- 0L
```

## The map

```{r plot}
show_classes(m, cls)
```

This is Fig. 1 of the paper, cell for cell, all 1178 of them.

## The seven layers

The map above is one view of the result. The other is the seven layers it was
assembled from, one per class. They are drawn after the order of the
assembling step, so each cell appears in one of them only.

```{r composition, echo = FALSE, fig.height = 8}
names7 <- c(core = "core", edge = "edge", perforation = "perforation",
            islet = "islet", bridge = "bridge", loop = "loop",
            branch = "branch")
layers7 <- lapply(names(names7), function(nm) as.numeric(cls == classes[[nm]]))
seven <- do.call(msk_add, c(list(msk_select(mspa, pattern)),
                            setNames(layers7, names(names7))))
specs <- lapply(names(names7), function(nm){
  .layer(nm, colours = c("#eeeeee", key$colour[key$id == classes[[nm]]]),
         legend = FALSE)
})
do.call(msk_vis, c(list(seven), specs))
```

The seven layers are not a division of the habitat on their own. Eighteen
cells of this pattern meet two definitions at once: eleven are both edge and
bridge, seven both perforation and loop. That is why the assembling step needs
an order.

## The whole classification in one function

The steps above, collected into one function. It takes a mosaik with a binary
layer `pattern` and returns the class code of every cell. The border width is
fixed at `size = 1` and written out as `sqrt(2)`: `mdf_loop()` runs the
recipes `peel` and `grow`, and a recipe carries its values with it instead of
reading variables around it.

```{r classify}
mspa_classify <- function(m) {
  m <- mdf_distance(m, source = "background", layer = "pattern", add = "edt") |>
    mdf_filter(edt > sqrt(2), add = "core") |>
    mdf_componentise(connectivity = 8L, layer = "pattern", add = "cc")
  m <- mdf_filter(m, pattern != 0, value = TRUE, layer = "cc") |>
    mdf_replace(old = NA, new = 0, layer = "cc") |>
    mdf_summarise(by = "cc", fun = "any", layer = "core", background = 0,
                  add = "hasCore") |>
    mdf_filter(pattern == 1 & hasCore == 0, add = "islet") |>
    mdf_distance(source = "foreground", layer = "core", add = "dcore") |>
    mdf_filter(core == 1 | (dcore > 0 & dcore <= sqrt(2)), add = "opening") |>
    mdf_offset(value = 0, layer = "core", add = "work") |>
    mdf_fill(connectivity = 4L, layer = "work", add = "corefill") |>
    mdf_filter(corefill == 1 & work == 0, add = "holes") |>
    mdf_filter(pattern < 0, add = "edge")
  peel <- mdf_distance(source = "foreground", layer = "corefill", add = "dfill") |>
    mdf_filter(edge == 1 | (corefill == 0 & dfill > 0 & dfill <= sqrt(2)),
               add = "edge") |>
    mdf_fill(connectivity = 8L, layer = "holes", add = "i0") |>
    mdf_filter(corefill == 1 & i0 == 0, add = "corefill") |>
    mdf_filter(work == 1 & corefill == 0, add = "work") |>
    mdf_fill(connectivity = 4L, layer = "work", add = "corefill") |>
    mdf_filter(corefill == 1 & work == 0, add = "holes")
  if (sum(msk_pull(m, "corefill")) > 0) {
    m <- mdf_loop(m, peel, until = sum(corefill, na.rm = TRUE) == 0)
  }
  m <- mdf_filter(m, edge == 1 & core == 0, add = "edge") |>
    mdf_filter(core == 0 & dcore > 0 & dcore <= sqrt(2) & edge == 0,
               add = "perforation") |>
    mdf_filter(pattern == 1 & core == 0 & islet == 0 & perforation == 0 &
                 edge == 0, add = "residues") |>
    mdf_filter(opening == 1 | residues == 1, add = "skin") |>
    mdf_skeletonise(anchor = "core", background = 0, layer = "skin",
                    add = "sk", method = "homotopic") |>
    mdf_filter(sk == 1 & core == 0, add = "sk") |>
    mdf_filter((opening == 1 & core == 0) | residues == 1, add = "i0") |>
    mdf_offset(value = 0, layer = "sk", add = "rec")
  grow <- mdf_dilate(struct = msk_struct("square", width = 3), layer = "rec") |>
    mdf_filter(rec == 1 & i0 == 1, add = "rec")
  m <- mdf_loop(m, grow, stable = TRUE, layer = "rec") |>
    mdf_filter(sk == 1 & i0 == 1 & rec == 1, add = "connector") |>
    mdf_componentise(connectivity = 8L, layer = "connector", add = "strand")
  m <- mdf_filter(m, connector != 0, value = TRUE, layer = "strand") |>
    mdf_replace(old = NA, new = 0, layer = "strand") |>
    mdf_componentise(connectivity = 8L, layer = "core", add = "coreID")
  m <- mdf_filter(m, core != 0, value = TRUE, layer = "coreID") |>
    mdf_replace(old = NA, new = 0, layer = "coreID") |>
    mdf_summarise(by = "strand", fun = "n_distinct", neighbours = TRUE,
                  connectivity = 8L, layer = "coreID", background = 0,
                  add = "nCores") |>
    mdf_summarise(by = "strand", fun = "n", layer = "strand", background = 0,
                  add = "strandSize") |>
    mdf_filter(connector == 1 & nCores >= 2 & strandSize >= 2, add = "bridge") |>
    mdf_filter(connector == 1 & bridge == 0, add = "loop") |>
    mdf_filter(pattern == 1 & core == 0 & islet == 0 & perforation == 0 &
                 edge == 0 & bridge == 0 & loop == 0, add = "branch")
  cls <- rep(0L, msk_ncells(m))
  for (nm in names(classes)) cls[msk_pull(m, nm) == 1] <- classes[[nm]]
  cls[msk_pull(m, "pattern") != 1] <- 0L
  cls
}
```

To check it, the function runs on the pattern again, and its class codes are
compared with `cls`, the codes built step by step above, which match Fig. 1 of
the paper. The comparison counts the cells where both agree. The pattern has
`r msk_ncells(mspa)` cells, so a count of `r msk_ncells(mspa)` means the
function gives the published class on every cell:

```{r classify-check}
same <- mspa_classify(mspa) == cls
sum(same)
```

## On an ordinary map

The pattern of the paper was drawn to show the method: every class appears,
cleanly, where one expects it. The forest of `landscape` is an ordinary map by
comparison, with irregular outlines, a road through a patch, a glade and a
windthrow, hedgerows, and a corridor cut by a river. Both maps are read at
`size = 1`, and the patches of `landscape` are, if anything, the larger.

```{r landscape}
forest <- mdf_filter(landscape, cover == 47, add = "pattern")
lcls <- mspa_classify(forest)
table(factor(lcls, c(1, 6, 3, 4, 5, 2, 7),
             c("core", "edge", "perforation", "islet", "bridge", "loop",
               "branch")))
```

```{r landscape-fig, echo = FALSE, fig.width = 7, fig.height = 6.5}
show_classes(forest, lcls)
```

Much of this reads well. The core is the interior of the large patches, and
the hedgerows and small patches are islets. The one bridge is the corridor
that joins the forest in the south-west to the rest. The corridor between the
two large northern patches, cut by the river, falls apart into two branches,
as it should.

Three things read less well. They follow from the definitions, not from this
map.

- **Corners become branches.** Of the 84 branch cells, 61 are the dead ends
  one would call so: the two halves of the cut corridor, and the strip that
  leaves the north-eastern patch to the south. The other 23, in ten pieces of
  one to nine cells, are corners and spikes of large patches. Whatever lies
  more than `sqrt(2)` cells from the core and connects no two core areas is
  branch, whether it leads somewhere or is just an irregular outline.
- **Edge exists only next to core.** Of the outermost forest cells of the
  patches that hold core, 216 of 357 are edge. The others are branch, loop,
  bridge or perforation. The edge of MSPA is the border of the core, not the
  border of the forest.
- **A perforation needs a hole that core surrounds on every side.** The glade
  in the north-eastern patch lies deep enough and is ringed by 12 cells of
  perforation. The windthrow in the north-western patch lies just as much
  inside the forest, but so close to its outline that no closed ring of core
  forms around it, and its border is edge.

None of this is an error of the implementation. The classification above
reproduces the published one cell for cell. The reference implementation of
MSPA (`mspa_win64.exe`, shipped with GuidosToolbox, run at `size = 1` with
8-connected habitat) gives the same classes on this map for 3359 of its 3360
cells, the corners and spikes included. The one cell that differs is the
bottom right corner of the map, where the two treat the border of the map
differently. It is what seven geometric definitions do to an outline that was
not drawn for them. A larger `size` moves the threshold but not the logic:
whatever lies more than `size` times `sqrt(2)` cells beyond the core is still
branch, loop or bridge.

## What to take from this

No part of this needed a function written for MSPA. The whole classification is
eight operators (`mdf_distance`, `mdf_filter`, `mdf_componentise`,
`mdf_replace`, `mdf_summarise`, `mdf_fill`, `mdf_dilate`, `mdf_skeletonise`)
and `mdf_loop()`, which repeats a recipe until a condition holds.

That is what mosaik is built for. It provides the operators that named methods
are made of, so the same parts make FRAGSTATS metrics, neutral landscapes and
measures of connectivity. Because every step is written out, you can read what
a method does and change it.

Two things are worth knowing if you adapt this:

- Only 8-connected habitat (`conn = 8L`) is covered. The 4-connected variant
  needs a different rule in the thinning step, which is not implemented.
- A `size` above 1 widens the core threshold. The step that finds the
  connecting strips would then need wider strips as well, which this vignette
  does not do.

## Reference

Soille, P. & Vogt, P. (2009). Morphological segmentation of binary patterns.
*Pattern Recognition Letters* 30(4), 456–459.

Soille, P. & Vogt, P. (2022). Morphological segmentation of binary patterns:
the open source release. *ISPRS Archives* XLVIII-4/W1, 427–432.
