---
title: "Morphological spatial pattern analysis, built from operators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphological spatial pattern analysis, built from operators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 6)
```

## What this vignette shows

Morphological spatial pattern analysis (MSPA) partitions the foreground of a
binary map (habitat against matrix) into seven mutually exclusive
morphological classes. It describes configuration rather than composition:
which parts of the habitat function as interior, as ecotone, and as structural
connection.

mosaik has no MSPA function. It does not need one. MSPA decomposes into a
handful of ordinary operations (a distance transform, a threshold, connected
components, a zonal summary, a thinning) which mosaik already provides. What
follows composes the full classification from those parts and reproduces the
published result cell for cell.

If you want the classification and not the explanation, skip to the last
section and take the code.

## The seven classes

Everything is defined relative to the **core**: foreground cells lying further
than a chosen distance from the matrix. That distance, `size`, is the method's
only real parameter, and it sets the depth at which interior habitat is
considered buffered from edge influence.

The remaining six classes describe how the non-core foreground relates to it:

| class | definition | reading |
|---|---|---|
| core | interior beyond `size` of the matrix | buffered habitat |
| edge | the outer boundary of a core area | the ecotone facing the matrix |
| perforation | the boundary of a hole enclosed within a core area | internal ecotone |
| islet | a patch containing no core at all | too small to hold interior; a stepping stone |
| bridge | a connector joining **two different** core areas | corridor |
| loop | a connector returning to the same core area | detour, not a corridor |
| branch | a connector attached at one end only | dead end |

The bridge/loop distinction carries most of the connectivity signal: a bridge
links two habitat blocks and its loss fragments them, whereas a loop returns
where it came from and its loss does not.

## Setup

The `mspa` dataset that ships with the package is the pattern from the paper
which introduced the method (Soille & Vogt 2009), in its layer `pattern`. That
paper prints the resulting classification as its Fig. 1, so the map produced at
the end of this vignette can be compared against it directly.

```{r setup, message = FALSE}
library(mosaik)

m <- mspa
ncols <- msk_dims(mspa)[1]; nrows <- msk_dims(mspa)[2]
```

The binary input is the layer `pattern`, and every layer we derive is added
alongside it, so the object carries the whole working state as we go. This is
what we start from: habitat in black, matrix in white.

```{r input, echo = FALSE}
msk_vis(m, .layer("pattern", colours = c("#ffffff", "#222222")))
```

Two settings. `size` is the edge width, meaning the depth at which habitat
counts as interior. `edu` is the unit it is measured in. The reference
implementation fixes this at `sqrt(2)`, the spacing of diagonal neighbours, so
`size = 1` means "further than `sqrt(2)` cell widths from the matrix".

```{r params}
size <- 1
edu  <- sqrt(2)

conn   <- 8L   # cells touching at a corner count as connected (habitat)
connBg <- 4L   # ... but not for the background
```

Throughout, each class is stored as its own 0/1 layer, and `mdf_filter()` is
what writes them: give it a condition over the layers, and it writes 1 where
the condition holds and 0 elsewhere.

Each class has a code and a colour, taken from the paper so the maps here can be
held against its Fig. 1. Code 8 is not a class of the method: it marks habitat
not yet classified, drawn grey, and it is how the figures below show what is
still to be explained.

```{r palette}
classes <- c(branch = 7, edges = 6, perf = 3, patch = 4,
             core = 1, bridge = 5, loop = 2)

mspaCols <- c("#ffffff",   # 0 background
              "#399c39",   # 1 core
              "#f7ff05",   # 2 loop
              "#0401ed",   # 3 perforation
              "#9a3312",   # 4 islet
              "#f10606",   # 5 bridge
              "#010100",   # 6 edge
              "#efce09",   # 7 branch
              "#cccccc")   # 8 not yet classified
```

The map after any number of steps is drawn by `show_state()`, which paints the
classes decided so far and leaves the rest of the habitat grey. As the
classification proceeds the grey shrinks.

```{r show-state}
show_state <- function(obj, done = character()){
  cls <- rep(0L, nrows * ncols)
  cls[msk_pull(obj, "pattern") == 1] <- 8L
  for(nm in done) cls[msk_pull(obj, nm) == 1] <- as.integer(classes[[nm]])

  state <- mosaik(extent = c(0, ncols, 0, nrows), res = 1,
                  vals = list(mspa = as.numeric(cls)))
  msk_vis(state, .layer("mspa", colours = mspaCols[sort(unique(cls)) + 1L]))
}
```

## Core

The distance transform gives every foreground cell its distance to the nearest
matrix cell. Core is what survives the threshold.

```{r core}
m <- mdf_distance(m, source = "background", layer = "pattern", add = "edt")
m <- mdf_filter(m, edt > size * edu, add = "core")
```

The green is core. Everything still grey is within `size` of the matrix, or
belongs to a patch too narrow to hold any interior at all, and the rest of the
method is about telling those cases apart.

```{r core-fig, echo = FALSE}
show_state(m, "core")
```

## Islet

An islet is a patch holding no core. Label the patches with
`mdf_componentise()`, then ask of each whether any of its cells is core.

`mdf_summarise()` does the second part. It groups cells by a label layer and writes
one summary value back onto every cell of the group, which is the raster
equivalent of a group-by. Here the summary is whether the patch contains a core
cell, and the patches answering no are the islets.

```{r islet}
m <- mdf_componentise(m, connectivity = conn, layer = "pattern", add = "cc")
m <- mdf_filter(m, pattern != 0, value = TRUE, layer = "cc") |>
    mdf_replace(old = NA, new = 0, layer = "cc")
m <- mdf_summarise(m, by = "cc", fun = "any", layer = "core",
               background = 0, add = "hasCore")
m <- mdf_filter(m, pattern == 1 & hasCore == 0, add = "patch")
```

The brown patches are gone from further consideration: having no core, they can
carry no boundary and connect nothing.

```{r islet-fig, echo = FALSE}
show_state(m, c("core", "patch"))
```

## Edge and perforation

Both are the one-cell boundary of a core area, and they differ in which side
they face. An **edge** faces the surrounding matrix, a **perforation** faces a
hole enclosed inside the core area.

Separating them is not a single test, because a patch may sit inside a hole of
another patch, whose own boundary is then an edge again. The reference handles
this by peeling: take the boundary of what you have, discard the part lying
inside a hole, and repeat, so each nested level is resolved in turn.

The starting state. Filling the holes of the core gives `corefill`, and what the
filling added is `holes`, the holes themselves.

```{r edges-init}
m <- mdf_distance(m, source = "foreground", layer = "core", add = "dcore")
m <- mdf_filter(m, core == 1 | (dcore > 0 & dcore <= size * edu), add = "opening")

m <- mdf_offset(m, value = 0, layer = "core", add = "work")
m <- mdf_fill(m, connectivity = connBg,
              layer = "work", add = "corefill")
m <- mdf_filter(m, corefill == 1 & work == 0, add = "holes")

# an empty accumulator: no cell satisfies pattern < 0, so this is all zeros
m <- mdf_filter(m, pattern < 0, add = "edges")
```

One round of peeling adds the boundary of the current `corefill` to `edges`,
then strips away the part enclosed in holes so the next round sees the next
level down.
Written with `obj = NULL`, a chain of `mdf_*` calls records itself instead of
executing. That recorded sequence is a **recipe**, and `mdf_loop()` replays it
until a condition is met.

```{r edges-loop}
peel <- mdf_distance(source = "foreground", layer = "corefill", add = "dfill") |>
  mdf_filter(edges == 1 | (corefill == 0 & dfill > 0 & dfill <= size * edu), add = "edges") |>
  mdf_fill(connectivity = conn,
           layer = "holes", add = "i0") |>
  mdf_filter(corefill == 1 & i0 == 0,
             add = "corefill") |>
  mdf_filter(work == 1 & corefill == 0,
             add = "work") |>
  mdf_fill(connectivity = connBg,
           layer = "work", add = "corefill") |>
  mdf_filter(corefill == 1 & work == 0,
             add = "holes")
```

```{r edges-converge}
if(sum(msk_pull(m, "corefill")) > 0){
  m <- mdf_loop(m, peel, until = sum(corefill, na.rm = TRUE) == 0)
}
```

`mdf_loop()` evaluates its condition after each round, so the `if` guards the
degenerate case of a map with no core at all, where no round should run.

How many rounds this takes depends on how deeply the pattern nests. Each round
resolves one level, and on this pattern the first round already collects all
`r sum(msk_pull(m, "edges"))` boundary cells, because nothing here sits inside
the hole of something else. A pattern with a patch inside a clearing inside a
forest block would need the rounds that follow.

Now separate the two. Everything the rounds collected, minus the core itself,
is edge. Perforation is the remainder of the core's boundary, the part the
peeling never reached because it faces inward.

```{r edges-split}
m <- mdf_filter(m, edges == 1 & core == 0, add = "edges")
m <- mdf_filter(m, core == 0 & dcore > 0 & dcore <= size * edu & edges == 0, add = "perf")
```

Black is edge, blue perforation. The core areas are now fully bounded, and what
stays grey touches no core area on any side: it lies between them.

```{r edges-fig, echo = FALSE}
show_state(m, c("core", "patch", "edges", "perf"))
```

## Connectors

What remains unclassified is connective tissue of some kind.

```{r residues}
m <- mdf_filter(m, pattern == 1 & core == 0 & patch == 0 & perf == 0 & edges == 0, add = "residues")
```

To find which of it genuinely connects, reduce the pattern to its centre lines
while holding the core areas fixed. That is `mdf_skeletonise(anchor = "core")`
with homotopic thinning, which removes a cell only where doing so preserves the
topology: it may not split a component or open a hole.

The consequence is what makes the classification work. A strand joining two core
areas is topologically necessary, so it survives thinning. A dead-end strand is
necessary to nothing, so
it retracts entirely. The branch/connector distinction falls out of the
thinning, so no separate test is needed for it.

```{r skeleton}
m <- mdf_filter(m, opening == 1 | residues == 1, add = "skin")
m <- mdf_skeletonise(m, anchor = "core", background = 0, layer = "skin",
                     add = "sk", method = "homotopic")
m <- mdf_filter(m, sk == 1 & core == 0, add = "sk")
```

Keep the parts of the skeleton reachable through the unclassified cells. Growing
a layer until it stops changing, a fixpoint, is `mdf_loop(stable = TRUE)`.

```{r connector}
m <- mdf_filter(m, (opening == 1 & core == 0) | residues == 1, add = "i0")

m <- mdf_offset(m, value = 0, layer = "sk", add = "rec")
grow <- mdf_dilate(struct = msk_struct("square", width = 3), layer = "rec") |>
        mdf_filter(rec == 1 & i0 == 1, add = "rec")
m <- mdf_loop(m, grow, stable = TRUE, layer = "rec")

m <- mdf_filter(m, sk == 1 & i0 == 1 & rec == 1, add = "connector")
```

What survived is drawn here over the grey it was cut from. These strands are the
connective structure of the pattern, and all that is left is to say what each
one connects.

```{r connector-fig, echo = FALSE}
conCls <- rep(0L, nrows * ncols)
conCls[msk_pull(m, "pattern") == 1] <- 8L
conCls[msk_pull(m, "connector") == 1] <- 5L
conState <- mosaik(extent = c(0, ncols, 0, nrows), res = 1,
                   vals = list(mspa = as.numeric(conCls)))
msk_vis(conState,
        .layer("mspa", colours = mspaCols[sort(unique(conCls)) + 1L]))
```

## Bridge, loop and branch

Label each connector and each core area, then ask two things of every connector:
how many distinct core areas it adjoins, and how many cells it spans. Both are
`mdf_summarise()` summaries, the first taken over the neighbours of each connector
rather than over its own cells.

A connector adjoining two or more core areas is a **bridge**, provided it is
longer than one cell. That second condition matters: a single cell can sit
diagonally in the gap between two core areas, adjoining both without running
between them, and the reference classifies it as a loop.

```{r bridge-loop}
m <- mdf_componentise(m, connectivity = conn, layer = "connector", add = "strand")
m <- mdf_filter(m, connector != 0, value = TRUE, layer = "strand") |>
    mdf_replace(old = NA, new = 0, layer = "strand")
m <- mdf_componentise(m, connectivity = conn, layer = "core", add = "coreID")
m <- mdf_filter(m, core != 0, value = TRUE, layer = "coreID") |>
    mdf_replace(old = NA, new = 0, layer = "coreID")

m <- mdf_summarise(m, by = "strand", fun = "n_distinct", neighbours = TRUE,
               connectivity = conn, layer = "coreID", background = 0,
               add = "nCores")
m <- mdf_summarise(m, by = "strand", fun = "n", layer = "strand",
               background = 0, add = "strandSize")

m <- mdf_filter(m, connector == 1 & nCores >= 2 & strandSize >= 2, add = "bridge")
m <- mdf_filter(m, connector == 1 & bridge == 0, add = "loop")
```

Red is bridge, yellow loop. This is the distinction the method exists to draw:
losing a red strand separates two habitat blocks, losing a yellow one does not.

```{r bridge-loop-fig, echo = FALSE}
show_state(m, c("core", "patch", "edges", "perf", "bridge", "loop"))
```

Branch is the residue, being what the thinning retracted because it leads
nowhere.

```{r branch}
m <- mdf_filter(m, pattern == 1 & core == 0 & patch == 0 & perf == 0 & edges == 0 &
                   bridge == 0 & loop == 0, add = "branch")
```

No grey is left, so every foreground cell now holds a class and the
classification is complete.

```{r branch-fig, echo = FALSE}
show_state(m, names(classes))
```

## Assembling

Collapse the seven layers into one map of class codes. Order matters where
classes overlap, since a cell can be both boundary and connector. The connector
takes precedence, so bridge and loop are written last.

```{r assemble}
cls <- rep(0L, nrows * ncols)
for(nm in names(classes)) cls[msk_pull(m, nm) == 1] <- as.integer(classes[[nm]])
cls[msk_pull(m, "pattern") != 1] <- 0L
```

## The map

```{r plot}
res <- mosaik(extent = c(0, ncols, 0, nrows), res = 1,
              vals = list(mspa = as.numeric(cls)))
msk_vis(res, .layer("mspa", colours = mspaCols[sort(unique(cls)) + 1L]))
```

This agrees with Fig. 1 of the paper cell for cell, all 1178 of them.

## The seven layers

The map above is one view of the result. The other is the seven binary layers it
was assembled from, each holding one class, which is what the pipeline actually
produced.

```{r composition, echo = FALSE, fig.height = 8}
op <- par(mfrow = c(4, 2), mar = c(0.5, 0.5, 2, 0.5))
labels <- c(core = "core", edges = "edge", perf = "perforation",
            patch = "islet", bridge = "bridge", loop = "loop",
            branch = "branch")
for(nm in names(labels)){
  # drawn after the precedence in "Assembling", so the 18 cells claimed by two
  # layers appear only in the one that wins
  v <- matrix(as.integer(cls == classes[[nm]]),
              nrow = nrows, ncol = ncols, byrow = TRUE)
  image(t(v[nrows:1, ]), col = c("#eeeeee", mspaCols[classes[[nm]] + 1L]),
        axes = FALSE, asp = nrows / ncols, main = labels[[nm]])
}
par(op)
```

The seven do not quite fall out as a partition on their own. Eighteen cells of
this pattern satisfy two definitions at once, eleven of them both edge and
bridge, seven both perforation and loop, which is why the assembling step needs
an order at all. The panels above apply that order, so each cell appears once.

## What to take from this

No part of this needed a function written for MSPA. The whole classification is
eight operators (`mdf_distance`, `mdf_filter`, `mdf_componentise`, `mdf_replace`,
`mdf_summarise`, `mdf_fill`, `mdf_dilate`, `mdf_skeletonise`) plus `mdf_loop()` to
iterate a recipe to a condition.

This is the design intent of the package. It provides the operators that named
methods decompose into, so the same parts compose FRAGSTATS-style metrics,
neutral models and connectivity measures. Because the composition is explicit,
you can read what a method does and change it.

Two things are worth knowing if you adapt this:

- Only 8-connected habitat (`conn = 8L`) is covered here. The 4-connected
  variant needs a different rule inside the thinning step and is not
  implemented.
- `size` above 1 widens the core threshold; the connector step would then also
  need the strands widened to match, which this vignette does not do.

## Reference

Soille, P. & Vogt, P. (2009). Morphological segmentation of binary patterns.
*Pattern Recognition Letters* 30(4), 456–459.

Soille, P. & Vogt, P. (2022). Morphological segmentation of binary patterns:
the open source release. *ISPRS Archives* XLVIII-4/W1, 427–432.
