---
title: "Pipelines: define once, apply to many rasters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipelines: define once, apply to many rasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r}
library(mosaik)
```

Every `mdf_*` operation records itself in the object's provenance. The same
record can be made without any data: called without an input object, an
`mdf_*` function does not run, it adds itself to a list of steps. That list is
a **recipe**, and `mdf()` applies it to a real mosaik. An algorithm defined this
way is written once and applied to any number of rasters, and it carries its own
description of what it does.

## A recipe

The core of a class is what is left after eroding it by one cell, the part that
is not edge. As a recipe, for class 47 of a layer called `cover`:

```{r}
core <- mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_erode(layer = "forest", add = "core")
```

Nothing has been computed yet. `mdf()` runs the steps on a real object:

```{r}
result <- mdf(landscape, core)
msk_names(result)
```

The same recipe applies to any raster that has a `cover` layer, here the example
data rotated by 90 degrees:

```{r}
rotated <- mdf(mdf_rotate(landscape, angle = 90), core)
sum(msk_pull(rotated, "core") == 1)
```

## Branching through named layers

A step reads what an earlier step wrote by its layer name. That is how a
pipeline branches and joins again, without loops or temporary variables. The
holes inside the patches of a class are the cells that filling adds:

```{r}
holes <- mdf_binarise(match = 41, layer = "cover", add = "class41") |>
  mdf_fill(layer = "class41", add = "filled") |>
  mdf_blend(layers = c("filled", "class41"), fun = "-", add = "holes")

h <- mdf(landscape, holes)
sum(msk_pull(h, "holes") == 1, na.rm = TRUE)
```

```{r, fig.alt = "Class 41 and the holes enclosed by its patches"}
msk_vis(h, .layer("class41"), .layer("holes"))
```

## Measuring what a pipeline produced

Recipes hold `mdf_*` steps. The measurements come after, on the result, so the
full sequence is: apply the recipe, then measure its layers.

```{r}
result |>
  msr_area(scale = "class", layer = "core") |>
  msk_categories()
```

## Repeating a recipe

Some algorithms repeat a step until nothing changes. `mdf_loop()` runs a recipe
either a fixed number of times or until a condition holds. Here a single forest
cell spreads through the forest until it stops growing; what is left is the
forest patch that cell belongs to:

```{r}
m <- mdf_binarise(landscape, match = 47, layer = "cover", add = "forest")
seed <- as.numeric(seq_len(msk_ncells(m)) ==
                   which(msk_pull(m, "forest") == 1)[1])
m <- msk_set(m, "seed", seed, prov = msk_prov("seed", list()))

grow <- mdf_dilate(struct = msk_struct("square", width = 3), layer = "seed") |>
  mdf_filter(seed == 1 & forest == 1, value = TRUE, background = 0,
             add = "seed")

m <- mdf_loop(m, grow, stable = TRUE, layer = "seed")
sum(msk_pull(m, "seed"))
```

```{r, fig.alt = "The forest and the one patch reached from the seed cell"}
msk_vis(m, .layer("forest"), .layer("seed"))
```

## Every step is on record

The steps a result went through are in its provenance, in order:

```{r}
unlist(lapply(msk_provenance(h), names))
```

The MSPA vignette takes this furthest: a complete published classification
algorithm, Morphological Spatial Pattern Analysis, built from eight operators
and `mdf_loop()`, matching the original cell for cell
(`vignette("mspa")`).
