---
title: "Techniques for combining operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Techniques for combining operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3)
```

```{r}
library(mosaik)
```

Each operation in mosaik does one small thing. What makes them useful is how
they combine, and this vignette goes through the ways they do.

## Defining recipes

At some point the analysis done for one map has to be done for another one.
Usually that means digging out the old script and hoping it still runs.

In mosaik the steps can be written down first and run later. Any `mdf_*` or
`msr_*` function called without an object does not compute anything, it just
records itself. A chain of such calls is a recipe, and `mdf()` runs it on
whatever mosaik it is given. The core of a forest is the forest minus its edge:

```{r}
core <- mdf_filter(expr = cover == 47, add = "forest") |>
  mdf_erode(layer = "forest", add = "core")
```

At this point there is no map anywhere, so nothing has been computed. That
happens here:

```{r}
result <- mdf(landscape, core)
msk_names(result)
```

The recipe only needs a layer called `cover` (that has the class 47), so it runs on any map that has
that. For lack of a second example map, the recipe runs on the same map
rotated by 90 degrees. The rotation moves every cell, but each cell keeps its
neighbours, so the erosion removes the same forest cells, and both maps have
the same number of core cells:

```{r}
rotated <- mdf(mdf_rotate(landscape, angle = 90), core)
c(original = sum(msk_pull(result, "core") == 1),
  rotated  = sum(msk_pull(rotated, "core") == 1))
```

## Referring to any earlier layer

An ordinary R pipe hands each step only what the step before it returned. In
`x |> f() |> g() |> h()`, the function `h()` gets what `g()` returned; the
result of `f()` is gone unless it was saved in a variable first. To combine the
result of `f()` with the result of `g()`, the pipe has to be broken up into
temporary variables.

mosaik works differently on purpose. Every step writes its result as a layer
under the name given in `add`, and that layer stays in the object. Any later
step can name any earlier layer, so a recipe can branch and join again while
still being written as one pipe.

The holes in the patches of class 41 are the filled patches minus the original
patches. The last step subtracts `class41`, which was made two steps earlier,
from `filled`, which was made by the step before:

```{r}
holes <- mdf_filter(expr = cover == 41, add = "class41") |>
  mdf_fill(layer = "class41", add = "filled") |>
  mdf_blend(layers = c("filled", "class41"), fun = "-", add = "holes")

h <- mdf(landscape, holes)
```

```{r, fig.alt = "Class 41 and the holes enclosed by its patches"}
msk_vis(h, .layer("class41"), .layer("holes"))
```

## Measuring inside a recipe

Measurements can go into a recipe too, anywhere between the other steps. So a
recipe can measure something, change the map, and measure again. This one
measures the forest, erodes it to its core, measures the core, and works out
what share of the map the core takes up:

```{r}
core_share <- mdf_filter(expr = cover == 47, add = "forest") |>
  msr_area(layer = "forest") |>
  mdf_erode(layer = "forest", add = "core") |>
  msr_area(layer = "core") |>
  msr(equation = "area.self / sum(area.all) * 100", label = "pcore",
      layer = "core")

r <- mdf(landscape, core_share)
msk_categories(r, layer = "forest")$area
msk_categories(r, layer = "core")
```

```{r, include = FALSE}
forest_cells <- msk_categories(r, layer = "forest")$area[2]
core_cells <- msk_categories(r, layer = "core")$area[2]
core_pct <- round(msk_categories(r, layer = "core")$pcore[2], 1)
```

The forest has `r forest_cells` cells, its core `r core_cells` of them, and
the core covers `r core_pct` % of the map (class 1 of `core`; class 0 is
everything else). Every number
from the recipe is in the object, next to the layer it describes, and the
recipe runs unchanged on the next map.

An equation can also combine several layers. A name that is a layer reads
the cells of the class in focus: on the layer `forest`, `core.self` holds the
values of the layer `core` in the cells of the class in focus. Since `core` is
1 for core and 0 elsewhere, their sum is the core area of that class. The
share of the forest that is core:

```{r}
r <- msr(r, equation = "sum(core.self) / area.self * 100",
         label = "corepart", layer = "forest")
msk_categories(r, layer = "forest")$corepart
```

Just under half the forest is core (class 1). The same equation on a layer of
forest patches gives the core share of each patch, although eroding a patch
can leave several pieces of core or none: the cells of each patch are read,
whatever the core is made of.

## Metrics built on metrics

An equation in `msr()` can name any value stored in the object, and that
includes a metric computed by an earlier `msr()` call. So a metric can be
built from other metrics, as many levels deep as needed. Here the share of
each class (`pland`), then Shannon diversity from those shares, then evenness
from Shannon diversity and the number of classes:

```{r}
m <- landscape |>
  msr_area(layer = "cover") |>
  msr(equation = "area.self / sum(area.all) * 100", label = "pland",
      layer = "cover") |>
  msr(equation = "-sum(pland.all / 100 * log(pland.all / 100))",
      label = "shannon", layer = "cover") |>
  msr(equation = "shannon.all / log(length(pland.all))",
      label = "evenness", layer = "cover")

msk_global(m, layer = "cover")
```

Evenness is 1 when all classes cover the same share and falls towards 0 as
one class dominates. Written this way, each metric states what it is made of,
and a variation (a different reference, a weighting) changes one line.

A metric is computed once, when `msr()` runs. Measuring the area again
afterwards does not update `pland`, `shannon` or `evenness`; to recompute
them, run the chain again, or put it in a recipe.

## Reducing many values to one

Reducing means combining many values into one: a sum, a maximum, a mean, a
count. It happens in several places in mosaik, and the difference between them
is only which values are combined:

| over | operation | example |
|---|---|---|
| layers | `mdf_blend(fun = ...)` | the sum of several layers, cell by cell |
| a neighbourhood | `mdf_morph(merge = ...)` | the mean of each cell's surroundings |
| a zone | `mdf_summarise(fun = ...)` | the size of the patch a cell belongs to |
| classes or patches | `msr()` | `sum()` over classes in Shannon diversity |

`mdf_summarise()` is the one without a counterpart in most raster tools. It
reduces the values of a whole zone, for example a patch, and writes the result
back onto every cell of that zone. Here each forest cell gets the size of its
patch:

```{r}
z <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_summarise(by = "patch", fun = "n", layer = "patch", add = "size")
```

```{r, fig.alt = "Forest patches, each coloured by its size in cells"}
msk_vis(z, .layer("forest"), .layer("size"))
```

`by` names the zones and `layer` the values that are reduced within them.
Counting cells needs no values of its own, so here both are the patch layer.
With a different `layer`, the same step writes the core share of each forest
patch onto its cells, as a map rather than a table: the core cells summed
within each patch, divided by the patch size.

```{r}
z <- z |>
  mdf_erode(layer = "forest", add = "core") |>
  mdf_summarise(by = "patch", fun = "sum", layer = "core", add = "core_cells") |>
  mdf_blend(layers = c("core_cells", "size"), fun = "/", add = "core_share")
```

```{r, fig.alt = "Forest patches, each coloured by the share of it that is core"}
msk_vis(z, .layer("forest"), .layer("core_share"))
```

A neighbourhood operation cannot do this, because it does not know in advance
how far a patch reaches.

## Neighbourhood operations

Dilation, erosion and smoothing look like different operations, but they are
one: `mdf_morph()`. It visits every cell, lays a small pattern (the
structuring element, from `msk_struct()`) over the cell and its neighbours,
and computes a new value for the cell in two steps.

The structuring element is itself a small grid of values, for example 3 by 3
cells, whose centre sits on the cell being computed. Each of its cells
belongs to one neighbour. A 1 includes that neighbour; a cell left empty
excludes it, which is how a disc or a cross gets its shape. Values other than
1 act as weights. The two steps are:

1. **blend**: each covered value is combined with the matching value of the
   pattern, for example kept as it is (`identity`), tested for equality
   (`equal`) or multiplied (`product`);
2. **merge**: the blended values are reduced to one, for example with `max`,
   `min`, `mean` or `all`.

Choosing the two gives the familiar operations, and many others:

| blend | merge | result |
|---|---|---|
| `identity` | `max` | dilation: a cell becomes 1 if any neighbour is 1 |
| `identity` | `min` | erosion: a cell stays 1 only if all neighbours are 1 |
| `identity` | `mean` | smoothing: the average of the neighbourhood |
| `product` | `mean` | weighted smoothing, with the pattern as weights |

`mdf_dilate()` and `mdf_erode()` do what the first two rows do. The shape of
the neighbourhood is the pattern: a square, a disc, a diamond, a cross, or any
shape given by hand. Here the canopy height smoothed over a 5 by 5 square:

```{r, fig.alt = "The canopy height and its smoothed version"}
s <- landscape |>
  mdf_morph(struct = msk_struct("square", width = 5, height = 5),
            blend = "identity", merge = "mean", layer = "canopy",
            add = "smooth")
msk_vis(s, .layer("canopy"), .layer("smooth"))
```

## Repeating steps

Some algorithms repeat a step until nothing changes anymore. `mdf_loop()` does
that, either a set number of times or until the map stops changing.

Here one forest cell grows into its neighbours, but only where there is forest.
Once it has nowhere left to grow, it has covered the patch it started in:

```{r}
m <- mdf_filter(landscape, cover == 47, add = "forest")
m <- mdf_distance(m, source = "background", layer = "forest", add = "depth")
start <- which.max(msk_pull(m, "depth"))             # the deepest forest cell
seed <- as.numeric(seq_len(msk_ncells(m)) == start)
m <- msk_add(m, seed = seed)

grow <- mdf_dilate(struct = msk_struct("square", width = 3), layer = "seed") |>
  mdf_filter(seed == 1 & forest == 1,
             add = "seed")

m <- mdf_loop(m, grow, stable = TRUE, layer = "seed")
```

```{r, eval = FALSE}
msk_vis(m, .layer("forest"), .layer("seed"))
```

```{r, grow-animation, echo = FALSE, eval = requireNamespace("gifski", quietly = TRUE), fig.show = "animate", animation.hook = "gifski", interval = 0.15, fig.alt = "The seed spreading through its forest patch, one step at a time"}
# the figure of the call above, drawn as an animation: one frame per round of
# the loop, from the single seed cell to the whole patch, with the last frame
# held so the result can be seen
cur <- msk_add(mdf_filter(landscape, cover == 47, add = "forest"),
               seed = seed)
repeat {
  msk_vis(cur, .layer("forest"), .layer("seed"))
  nxt <- mdf(cur, grow)
  if (identical(msk_pull(nxt, "seed"), msk_pull(cur, "seed"))) break
  cur <- nxt
}
for (i in 1:10) msk_vis(cur, .layer("forest"), .layer("seed"))
```

```{r, echo = FALSE, eval = !requireNamespace("gifski", quietly = TRUE), fig.alt = "The forest and the one patch reached from the seed cell"}
# without gifski the figure is the final state only
msk_vis(m, .layer("forest"), .layer("seed"))
```

`mdf_componentise()` finds patches faster, and here it serves as a referee: if
the loop is right, the seed covers exactly the cells of the patch
`mdf_componentise()` finds around the starting cell. The square kernel reaches
diagonal neighbours, so the patches are counted with eight neighbours too:

```{r}
m <- mdf_componentise(m, connectivity = 8L, layer = "forest", add = "patches")
p <- msk_pull(m, "patches")
identical(which(msk_pull(m, "seed") == 1), which(p == p[start]))
```

But growing a seed can do what numbering patches cannot, because both the
start and the rule of growth are free:

- **the start**: seeds placed at chosen cells, say a nest site or where a river
  enters, find only the patches connected to those places;
- **the rule**: growth can be limited to any condition over the layers, not
  only "is forest", for example forest below a certain slope;
- **the distance**: stopping after `times = n` rounds gives everything within
  n steps of the seed when travelling only through forest, a distance that a
  straight line across a gap does not measure.

Grow, restrict, repeat is how many published methods work, and this is how
they are written in mosaik.

## Provenance: what was done

Every step records itself in the object it returns. Each entry holds the
function, all its arguments (defaults included), the layer it read, the layer
it wrote, the version of mosaik and the time. The entries use the vocabulary of
the W3C PROV standard for provenance. This is the entry for the fill step of
the holes recipe:

```{r}
str(msk_provenance(h)[[3]])
```

Printing the object lists the steps in short form:

```{r}
h
```

The record is not only documentation, it can be run again. The recipes at the
start of this vignette worked because a function called without an object
records itself instead of computing anything. A recipe is therefore a mosaik
that holds a history of steps and no data. A result such as `h` holds the same
kind of history, only with data next to it. `mdf()` reads nothing but the
history, so it accepts either one: the recipe `holes`, or any result made with
it.

That means an analysis does not have to be written as a recipe in advance.
Any result can serve as one later. Here the rotated map stands in for next
year's map, and it gets exactly the treatment that `h` received, taken from
`h` itself:

```{r}
next_map <- mdf_rotate(landscape, angle = 90)
again <- mdf(next_map, h)
msk_names(again)
```

Running the original recipe gives the same result:

```{r}
identical(msk_pull(again, "holes"),
          msk_pull(mdf(next_map, holes), "holes"))
```

One kind of step cannot be repeated: one whose input was too large to record
(over 10 kB), such as a whole layer of cell values given to `msk_add()`. The history keeps only a
fingerprint (digest) of such an input, which is enough to tell whether two
inputs were the same but not to rebuild it, and `mdf()` stops with an error
that names the step.

This matters wherever a number has to be defended. A figure in a report or a
monitoring series can be traced back to every step and argument that produced
it, a reviewer can rerun those steps, and the same method is applied to each
new map without anyone retyping it. A changed argument between two years shows
up in the record instead of hiding in a script.

`vignette("mspa")` takes all of this further and rebuilds Morphological
Spatial Pattern Analysis, a published classification method, from eight
operators and `mdf_loop()`.
