---
title: "What a patch borders"
summary: "The number of different land-cover classes around each patch, its edge contrast."
question: pressure
about: forest
scale: patch
declares: []
uses: [land-cover, patches, ecotone]
result:
  - {object: contrast, layer: classes, text: "its patch borders {v} land-cover classes"}
---

What lies around a patch acts on it. Counting the distinct classes along the
outside of each patch measures its edge contrast: a patch surrounded by one
kind of land borders one class, a patch between fields, roads and rivers many.

```{r}
library(mosaik)
contrast <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_summarise(by = "patch", fun = "n_distinct", neighbours = TRUE,
                layer = "cover", add = "classes")
```

```{r, fig.alt = "Forest patches, coloured by the number of classes they border"}
msk_vis(contrast, .layer("forest"), .layer("classes"))
```

**Check.** Patch 3, in the north-east, lies in one large meadow and touches
only the river and a fallow field besides; patches 4 and 5, in the west, lie
among many kinds of land:

```{r}
classes <- tapply(msk_pull(contrast, "classes"), msk_pull(contrast, "patch"),
                  unique)
classes[c("3", "4", "5")]
```

```{r, include = FALSE}
stopifnot(classes[["3"]] == 3, classes[["3"]] < min(classes[c("4", "5")]))
```
