---
title: "Ecotones"
summary: "The zone of transition between the forest and the land next to it."
question: arrangement
about: forest
scale: cell
declares: []
uses: [class, ecotone]
result:
  - {object: boundary, layer: boundary, text: "in the ecotone between forest and the land next to it"}
---

An ecotone is a zone of transition between adjacent ecological systems
(Holland 1988). On a map of cells, it reaches to both sides of the line where
two kinds of land meet. Growing the forest by one cell and subtracting the
forest shrunk by one cell leaves such a zone, two cells wide: its outer row
lies on the neighbouring land, its inner row in the forest. A wider ecotone is
a larger structuring element.

```{r}
library(mosaik)
boundary <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_dilate(layer = "forest", add = "grown") |>
  mdf_erode(layer = "forest", add = "shrunk") |>
  mdf_blend(layers = c("grown", "shrunk"), fun = "-", add = "boundary")
```

```{r, fig.alt = "The forest and the band along its boundary"}
msk_vis(boundary, .layer("forest"), .layer("boundary"))
```

**Check.** The inner row of the ecotone is exactly the forest that shrinking
removes:

```{r}
inner <- msk_pull(boundary, "boundary") == 1 & msk_pull(boundary, "forest") == 1
removed <- msk_pull(boundary, "forest") == 1 & msk_pull(boundary, "shrunk") == 0
identical(inner, removed)
```

```{r, include = FALSE}
stopifnot(identical(inner, removed))
```

**Reference.** Holland MM (1988) SCOPE/MAB technical consultations on
landscape boundaries: report of a SCOPE/MAB workshop on ecotones. Biology
International, Special Issue 17: 47-106.
