---
title: "Condition against a declared reference"
summary: "How far each forest patch is from a reference state, between a reference level and a collapse level."
question: state
about: forest
scale: patch
declares: ["the indicator", "the reference level", "the collapse level"]
uses: [canopy, declared, patches]
result:
  - {object: condition, layer: condition, text: "its patch scores {v} between collapse (0) and reference (1)", digits: 2}
---

A condition indicator says how far an ecosystem is from a reference state. A
common way to make indicators comparable is to rescale each one between a
reference level, which scores 1, and a lower level, which scores 0. Both
levels are declarations, and the recipe makes them two numbers in the code.

Here the indicator is the mean canopy height of each forest patch, a measure
of its structure. The reference is the height of an old forest, 25 m, and the
lower level a forest without canopy, 0 m:

```{r}
library(mosaik)
reference <- 25
collapse <- 0

condition <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_summarise(by = "patch", fun = "mean", layer = "canopy",
                add = "height") |>
  mdf_summarise(by = "patch", fun = "n", layer = "patch", add = "size") |>
  mdf_offset(fun = "-", value = collapse, layer = "height",
             add = "condition") |>
  mdf_offset(fun = "/", value = reference - collapse, layer = "condition")
```

The two `mdf_offset()` steps compute
`(height - collapse) / (reference - collapse)`, which is 1 at the reference
and 0 at the collapse level.

```{r, fig.alt = "The mean canopy height of each forest patch, and its condition"}
msk_vis(condition, .layer("height"), .layer("condition"))
```

**Check.** The canopy is lowest at the edge, and a small patch is mostly edge.
Small patches should therefore score lower than large ones, and no patch
should reach 1, since even the largest has an edge:

```{r, fig.alt = "The condition of each forest patch against its size", fig.width = 4, fig.height = 3}
patch <- msk_pull(condition, "patch")
size <- tapply(msk_pull(condition, "size"), patch, unique)
score <- tapply(msk_pull(condition, "condition"), patch, unique)
plot(size, score, log = "x", xlab = "patch size (cells)", ylab = "condition")
```

```{r, include = FALSE}
stopifnot(cor(log(size), score) > 0.5, max(score) < 1)
```

Size is not all of it: patch 5, in the south-west, is as large as patch 4 but
scores lower, because its stand is younger. A different declaration gives
different numbers. Taking the tallest stand of the region as the reference
would change every score, so a condition figure is only comparable with
others made against the same levels.
