---
title: "Connectivity through the land"
summary: "How hard it is to get from each patch of forest to the next, crossing the land as it is."
question: arrangement
about: forest
scale: patch
declares: ["the cost of crossing each kind of land"]
uses: [land-cover, patches, distance]
result:
  - {object: connectivity, layer: hardness, text: "getting from its patch to the nearest other is {v} times as hard as the straight distance suggests", digits: 1}
---

The straight distance between two patches says little about whether an animal
gets from one to the other. A road or a settlement in between is hard to
cross, a hedgerow or a meadow easy. A cost surface gives every kind of land a
cost of crossing it, and the cheapest path between two patches sums the costs
along the way.

The costs are a declaration and depend on the species. Here they are those of
an animal of the forest that crosses extensive grassland and fallow readily,
fields reluctantly, and roads, settlements and the river hardly at all:

```{r}
library(mosaik)
cost <- c("1" = 20, "11" = 5, "21" = 4, "24" = 2, "27" = 2, "31" = 30,
          "35" = 50, "41" = 3, "44" = 3, "47" = 1)

connectivity <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_replace(old = as.numeric(names(cost)), new = cost, layer = "cover",
              add = "friction") |>
  msr_cost(scale = "patch", layer = "forest") |>
  msr_cost(scale = "patch", cost = "friction", layer = "forest") |>
  msr(equation = "min(distance.patch)", label = "straight.patch", layer = "forest") |>
  msr(equation = "min(friction.patch)", label = "through.patch", layer = "forest")
p <- msk_patches(connectivity, layer = "forest")
```

For each patch, `straight` is the straight distance to the nearest other
patch, and `through` the cost of the cheapest path to the nearest other patch
across the land. Their ratio says how much harder the land makes the way:

```{r}
connectivity <- connectivity |>
  mdf_replace(old = p$patch, new = p$through / p$straight, layer = "patch",
              add = "hardness")
```

```{r, fig.alt = "The cost of crossing each cell, and for each patch how much harder the land makes the way to its neighbour"}
msk_vis(connectivity, .layer("friction"), .layer("hardness"))
```

**Check.** Crossing a cell costs at least 1, so no way can be cheaper than
the straight distance, and every ratio is at least 1. Many patches lie
exactly two cells from their nearest neighbour, one cell of open land
between them. If the land in between matters, their costs through the land
should differ, although their straight distances are the same:

```{r}
ratio <- data.frame(patch = p$patch, straight = round(p$straight, 1),
                    through = round(p$through, 1))
print(ratio[which(ratio$straight == 2), ], row.names = FALSE)
```

```{r, include = FALSE}
stopifnot(all(p$through >= p$straight, na.rm = TRUE),
          length(unique(ratio$through[which(ratio$straight == 2)])) > 2)
```

The recipe on isolation measures the straight distance only. Two patches
equally far apart can be well connected or cut off, and only the land in
between tells which.
