---
title: "Conversion of natural ecosystems"
summary: "Natural land at an earlier date that is no longer natural today."
question: change
about: forest
scale: landscape
declares: ["the forest at the earlier date", "what counts as natural"]
uses: [earlier, class]
result:
  - {object: conversion, layer: converted, text: "forest converted since the earlier date"}
---

Conversion compares two maps: one from an earlier date and one from today.
Natural land on the first that is not natural on the second has been
converted.

`landscape` holds only today's map, so the recipe first declares the forest of
the earlier date. The declaration is that patch 4, the large patch in the
north-west, was whole: the fallow and shrub inside the outline it has when
closed by three cells were forest, and so was the road where closing the
forest by one cell covers it. This is an assumption about the past, not an
observation, and the result holds only as far as the assumption does:

```{r}
library(mosaik)
square <- msk_struct("square", width = 3)
seven <- msk_struct("square", width = 7)
then_and_now <- landscape |>
  mdf_filter(cover == 47, add = "forest_now") |>
  mdf_componentise(connectivity = 8L, layer = "forest_now", add = "patch") |>
  mdf_filter(patch == 4, add = "four") |>
  mdf_dilate(struct = seven, layer = "four", add = "whole") |>
  mdf_erode(struct = seven, layer = "whole") |>
  mdf_dilate(struct = square, layer = "forest_now", add = "closed") |>
  mdf_erode(struct = square, layer = "closed") |>
  mdf_filter(forest_now == 1 | (whole == 1 & cover == 27) |
               (cover == 35 & closed == 1), add = "forest_then")
```

With a map of the earlier date at hand, the recipe starts here: the
conversion is what was forest then and is not now.

```{r}
conversion <- then_and_now |>
  mdf_filter(forest_then == 1 & forest_now == 0, add = "converted")
```

```{r, fig.alt = "The forest then, the forest now, and what was converted"}
msk_vis(conversion, .layer("forest_then"), .layer("forest_now"),
        .layer("converted"))
```

**Check.** The map was made with 39 cells of fresh clear-cut in patch 4, and
with the road crossing it. The converted cells should be these 39 cells, now
fallow and shrub (27), plus the road where it crosses the forest (35):

```{r}
converted <- table(msk_pull(conversion, "cover")[msk_pull(conversion, "converted") == 1])
converted
```

```{r, include = FALSE}
stopifnot(identical(names(converted), c("27", "35")), converted[["27"]] == 39)
```
