---
title: "Core"
summary: "The part of each patch far enough from its edge not to be affected by what lies outside."
question: arrangement
about: forest
scale: patch
declares: ["the edge width (cells)"]
uses: [distance, core]
result:
  - {object: core, layer: core, text: "in the core of its patch"}
  - {object: core, layer: depth, text: "{v} cells in from the forest edge", digits: 1, omit: 0}
---

The core of a patch is the part far enough from its edge not to be affected
by what lies outside. The distance of every forest cell to the nearest
non-forest cell, cut at a threshold, gives it. The threshold is the edge
width, here two cells:

```{r}
library(mosaik)
core <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_filter(depth >= 2, add = "core")
```

The distance is 1 in the outermost row of forest cells and grows towards the
inside of each patch; cells two or more cells from the edge are core:

```{r, fig.width = 9, fig.alt = "The forest, the distance of each forest cell to the edge, and the core"}
msk_vis(core, .layer("forest"),
        .layer("depth", bins = 11,
               colours = c("#f7fcf0", colorRampPalette(c("#78c679", "#004529"))(10))),
        .layer("core"))
```

**Check.** Hedgerows, islets and the thin strips are at most a few cells wide,
so none of the patches smaller than 12 cells should have any core. That needs
two numbers per patch, which `mdf_summarise()` gives: with `fun = "sum"` over
the core layer, the number of core cells in each patch, and with `fun = "n"`,
the number of cells of each patch, its size.

```{r}
s <- core |>
  mdf_summarise(by = "patch", fun = "sum", layer = "core", add = "core_cells") |>
  mdf_summarise(by = "patch", fun = "n", layer = "patch", add = "size")
```

```{r, fig.alt = "The number of core cells in each patch, and the size of each patch"}
msk_vis(s, .layer("core_cells"), .layer("size"))
```

The small patches are dark on both maps: few cells, and no core. Summed over
all patches smaller than 12 cells, the number of core cells is 0:

```{r}
small_core <- sum(msk_pull(s, "core_cells")[msk_pull(s, "size") < 12],
                  na.rm = TRUE)
small_core
```

```{r, include = FALSE}
stopifnot(small_core == 0)
```
