---
title: "Corridors"
summary: "Narrow strips of forest that join two cores, told apart from dead ends."
question: arrangement
about: forest
scale: patch
declares: ["the edge width (cells)"]
uses: [core, corridors]
result:
  - {object: corridors, layer: joins, text: "{label}", labels: {"0": "on a narrow strip without any core, such as a hedgerow", "1": "on a dead end: a narrow strip that leaves one core", "2": "in a corridor between two cores"}}
---

A corridor is a narrow strip of forest joining two cores. The skeleton of the
forest, its centre line, runs through cores and corridors alike; removing the
cores leaves the pieces of skeleton between them. A piece that touches two
different cores is a corridor, a piece that touches one is a dead end: a
branch, a loop that returns to its core, or a corridor that has been
interrupted.

The cores are the forest two or more cells from its edge:

```{r}
library(mosaik)
corridors <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_binarise(thresh = 2, layer = "depth", add = "core") |>
  mdf_componentise(connectivity = 8L, background = 0, layer = "core",
                   add = "core_id") |>
  mdf_skeletonise(layer = "forest", add = "skeleton") |>
  mdf_filter(skeleton == 1 & core == 0, add = "piece") |>
  mdf_componentise(connectivity = 8L, layer = "piece", add = "piece_id") |>
  mdf_summarise(by = "piece_id", fun = "n_distinct", neighbours = TRUE,
                layer = "core_id", add = "joins")
```

```{r, fig.alt = "Cores, and the skeleton pieces between them by the number of cores they join"}
msk_vis(corridors, .layer("core"), .layer("joins"))
```

**Check.** One corridor joins two cores: the one in patch 5, between the core
of its northern lobe and its round core in the south. The other pieces reach
one core, or none at all: those are the centre lines of hedgerows and islets,
which have no core of their own:

```{r}
joins <- tapply(msk_pull(corridors, "joins"), msk_pull(corridors, "piece_id"),
                unique)
table(joins)
```

```{r, include = FALSE}
stopifnot(sum(joins == 2) == 1, all(joins <= 2))
```
