---
title: "Critical places"
summary: "The narrow parts of the forest where a small clearing would split a patch in two."
question: arrangement
about: forest
scale: cell
declares: ["how narrow a part must be to count (cells)"]
uses: [patches, corridors]
result:
  - {object: critical, layer: cut_off, text: "a clearing here would cut off {v} cells of wide forest from the rest of its patch"}
---

A patch of forest holds together through its narrowest parts. A small
clearing there splits it into pieces that are each smaller and more isolated,
while the same clearing in a wide part only makes a hole. The recipe finds the
narrow parts that hold wide parts of the forest together.

Opening the forest, shrinking it and growing it again by one cell, removes
every part narrower than three cells and leaves the wide parts as they were.
The narrow parts are what opening removes:

```{r}
library(mosaik)
square <- msk_struct("square", width = 3)
critical <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_erode(struct = square, layer = "forest", add = "wide") |>
  mdf_dilate(struct = square, layer = "wide") |>
  mdf_filter(forest == 1 & wide == 0, add = "narrow")
```

Each wide part is numbered and measured, and each narrow part asks two things
of the wide parts next to it: how many different ones it touches, and how big
the smallest of them is. A narrow part touching two or more wide parts holds
them together, so a clearing in it would cut off at least the smallest:

```{r}
critical <- critical |>
  mdf_componentise(connectivity = 8L, layer = "wide", add = "wide_id") |>
  mdf_summarise(by = "wide_id", fun = "n", layer = "wide_id",
                add = "wide_size") |>
  mdf_componentise(connectivity = 8L, layer = "narrow", add = "narrow_id") |>
  mdf_summarise(by = "narrow_id", fun = "n_distinct", neighbours = TRUE,
                layer = "wide_id", add = "touches") |>
  mdf_summarise(by = "narrow_id", fun = "min", neighbours = TRUE,
                layer = "wide_size", add = "smallest") |>
  mdf_filter(touches >= 2, value = TRUE, layer = "smallest", add = "cut_off")
```

```{r, fig.alt = "The wide and the narrow parts of the forest, and how much wide forest a clearing in each critical narrow part would cut off"}
msk_vis(critical, .layer("wide"), .layer("narrow"), .layer("cut_off"))
```

**Check.** The corridor that joins the northern lobe of patch 5 to its round
core in the south is two cells wide, so it should be narrow, touch two wide
parts and be critical. A clearing in it would cut off the smaller of the two,
some tens of cells. The interior of the large patches is wide, so no critical
place should lie deeper than a few cells inside the forest:

```{r}
cut_off <- msk_pull(critical, "cut_off")
depth <- critical |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  msk_pull("depth")
# the cell in the corridor of patch 5: column 16, row 24 from the bottom
corridor <- (msk_dims(critical)[2] - 24) * msk_dims(critical)[1] + 16
c(critical_cells = sum(!is.na(cut_off)),
  in_the_corridor = cut_off[corridor],
  deeper_than_three = sum(!is.na(cut_off[depth > 3])))
```

```{r, include = FALSE}
stopifnot(cut_off[corridor] > 20, sum(!is.na(cut_off[depth > 3])) == 0)
```

The critical places are few, and each of them is where a new path, a ditch or
a building plot would do most harm.
