---
title: "Distance to disturbance"
summary: "How far each part of the forest lies from the road and the settlements."
question: pressure
about: forest
scale: cell
declares: ["what counts as a source of disturbance"]
uses: [class, distance]
result:
  - {object: disturbance, layer: from_disturbance, text: "{v} cells from the nearest road or settlement", digits: 1}
---

Noise, light, traffic and people reach into the forest from roads and
settlements. How far each part of the forest lies from them says how much of
it is quiet. Which land counts as a source of disturbance is a declaration;
here it is the road and the settlements.

```{r}
library(mosaik)
disturbance <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_binarise(match = c(31, 35), layer = "cover", add = "source") |>
  mdf_distance(layer = "source", add = "distance") |>
  mdf_filter(forest == 1, value = TRUE, layer = "distance",
             add = "from_disturbance")
```

```{r, fig.alt = "The road and the settlements, and the distance of each forest cell from them"}
msk_vis(disturbance, .layer("source"), .layer("from_disturbance"))
```

**Check.** The road runs through patch 4, and a settlement lies right beside
patch 5, so the forest of both should lie about equally close to
disturbance. Patch 3 lies east of the river, away from the road, and should
be the quietest of the three large patches. The quietest forest of all should
be the small patch in the south-east corner, which lies far from both:

```{r}
d <- msk_pull(disturbance, "from_disturbance")
patch <- msk_pull(mdf_componentise(disturbance, connectivity = 8L,
                                   layer = "forest", add = "patch"), "patch")
round(tapply(d, patch, median)[c("3", "4", "5")], 1)
patch[which.max(d)]
```

```{r, include = FALSE}
med <- tapply(d, patch, median)
stopifnot(abs(med[["4"]] - med[["5"]]) < 2,
          med[["3"]] > max(med[["4"]], med[["5"]]) + 3,
          patch[which.max(d)] == max(patch, na.rm = TRUE))
```

Core and distance to disturbance can disagree: a cell can be deep inside the
forest and still lie near a road that runs through it.
