---
title: "The edge-depth curve"
summary: "How far the effect of a forest edge reaches into the forest."
question: state
about: forest
scale: zone
declares: ["the width of a ring (cells)"]
uses: [canopy, distance, zones]
result:
  - {object: depth, layer: ring_mean, text: "the canopy this far into the forest stands {v} m tall on average", digits: 1}
---

At an old forest edge, light, wind and the land outside reach into the forest,
and the trees there grow lower than in the interior. The edge-depth curve
shows how far that effect reaches: the forest is divided into rings by their
distance from the edge, and the mean canopy height of each ring is plotted
against its depth.

```{r}
library(mosaik)
depth <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_filter(forest == 1, value = TRUE, layer = "depth") |>
  mdf_categorise(breaks = 1:8, layer = "depth", add = "ring") |>
  mdf_summarise(by = "ring", fun = "mean", layer = "canopy",
                add = "ring_mean")
```

`depth` is the distance of each forest cell to the nearest cell outside the
forest, so the outermost cells have depth 1. `mdf_categorise()` cuts it into
rings one cell wide, and `mdf_summarise()` writes the mean canopy height of
each ring onto its cells:

```{r, fig.alt = "The depth of each forest cell, its ring, and the mean canopy height of the ring"}
msk_vis(depth, .layer("depth"), .layer("ring"), .layer("ring_mean"))
```

The curve takes one value per ring:

```{r, fig.alt = "The mean canopy height of each ring against its depth", fig.width = 4, fig.height = 3}
curve <- tapply(msk_pull(depth, "ring_mean"), msk_pull(depth, "ring"), unique)
plot(as.numeric(names(curve)), curve, type = "b", xlab = "depth (cells)",
     ylab = "mean canopy height (m)")
```

**Check.** The canopy should rise over the first three rings and then level
off, so the edge effect reaches about three cells in. The deepest ring is
lower again: few cells lie that deep, and many of them belong to the younger
stand in the south-west, which is lower throughout:

```{r}
round(curve, 1)
```

```{r, include = FALSE}
stopifnot(all(diff(curve[1:3]) > 0), curve[3] - curve[1] > 5,
          abs(curve[4] - curve[3]) < 2)
```

The curve averages over all edges of the forest, old and new; the recipe
"Old and new edges" separates them.
