---
title: "Old and new edges"
summary: "Edges cut recently, which have not yet grown the lower mantle of old ones."
question: change
about: forest
scale: zone
declares: ["the forest at the earlier date"]
uses: [canopy, earlier, ecotone]
result:
  - {object: edges, layer: edge, text: "on {label}", labels: {"1": "an old forest edge", "2": "a new forest edge"}}
---

Over decades, a forest edge grows a mantle of shrubs and lower trees, so the
canopy rises gradually from the edge into the forest. A new edge has not had
that time: the trees that stood inside the forest stand at the edge at their
full height. An average over all edges mixes the two states.

A new edge is one that was not an edge at the earlier date. `landscape` holds
only today's map, so the forest of the earlier date is declared, as in the
recipe on conversion: patch 4 was whole, the fallow and shrub inside its
closed outline and the road through it were forest. The outermost row of
today's forest is then split into the cells that were edge already and those
that lay inside the forest then:

```{r}
library(mosaik)
square <- msk_struct("square", width = 3)
seven <- msk_struct("square", width = 7)
edges <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest_now") |>
  mdf_componentise(connectivity = 8L, layer = "forest_now", add = "patch") |>
  mdf_filter(patch == 4, add = "four") |>
  mdf_dilate(struct = seven, layer = "four", add = "whole") |>
  mdf_erode(struct = seven, layer = "whole") |>
  mdf_dilate(struct = square, layer = "forest_now", add = "closed") |>
  mdf_erode(struct = square, layer = "closed") |>
  mdf_filter(forest_now == 1 | (whole == 1 & cover == 27) |
               (cover == 35 & closed == 1), add = "forest_then") |>
  mdf_distance(source = "background", layer = "forest_now", add = "depth_now") |>
  mdf_distance(source = "background", layer = "forest_then", add = "depth_then") |>
  mdf_filter(forest_now == 1 & depth_now < 2, add = "rim") |>
  mdf_filter(rim == 1 & depth_then < 2, add = "old_edge") |>
  mdf_filter(rim == 1 & depth_then >= 2, add = "new_edge") |>
  mdf_blend(layers = c("old_edge", "new_edge"), fun = "+", weights = c(1, 2),
            add = "edge") |>
  mdf_filter(edge > 0, value = TRUE, layer = "edge") |>
  mdf_summarise(by = "edge", fun = "mean", layer = "canopy",
                add = "edge_mean")
```

```{r, fig.alt = "The canopy height, the old (1) and new (2) edges, and their mean canopy height"}
msk_vis(edges, .layer("canopy"), .layer("edge"), .layer("edge_mean"))
```

**Check.** The new edges run along the road and the fresh clear-cuts, where
the trees stand at full height. Their mean canopy should lie well above that
of the old edges, whose outermost trees belong to the mantle:

```{r}
means <- tapply(msk_pull(edges, "edge_mean"), msk_pull(edges, "edge"), unique)
round(means, 1)
```

```{r, include = FALSE}
stopifnot(means[["2"]] > means[["1"]] + 5)
```

An edge-depth curve over a forest with many new edges therefore shows less of
an edge effect than will develop along them.
