---
title: "Habitat within reach"
summary: "The share of forest within a given distance of every place on the map."
question: amount
about: forest
scale: cell
declares: ["the distance of reach (cells)"]
uses: [class, neighbourhood]
result:
  - {object: within, layer: share, text: "{v} % of the land within three cells is forest", percent: true}
---

A total area says how much forest there is, but not where it can be reached.
For an animal that moves a few hundred metres, or for people who walk to a
wood, what counts is the forest within reach of each place. The share of
forest within a disc around every cell gives it. The radius of the disc is the
reach, here three cells:

```{r}
library(mosaik)
reach <- 3
disc <- msk_struct("disc", width = 2 * reach + 1, height = 2 * reach + 1)
within <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_morph(struct = disc, blend = "product", merge = "mean",
            layer = "forest", add = "share")
```

```{r, fig.alt = "The forest, and the share of forest within three cells of each cell"}
msk_vis(within, .layer("forest"), .layer("share"))
```

**Check.** Deep inside the large patches, every cell within reach is forest,
so the share there should be 1. Far out in the open land, more than three
cells from the nearest forest, it should be 0:

```{r}
check <- within |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_distance(source = "foreground", layer = "forest", add = "away")
share <- msk_pull(check, "share")
c(deep = range(share[msk_pull(check, "depth") > reach + 1]),
  far = range(share[msk_pull(check, "away") > reach + 1]))
```

```{r, include = FALSE}
stopifnot(all(share[msk_pull(check, "depth") > reach + 1] == 1),
          all(share[msk_pull(check, "away") > reach + 1] == 0))
```

Between the two, the share falls off smoothly across the edge, and a hedgerow
lifts it in the open land around it.
