---
title: "Holes in patches"
summary: "The land a patch encloses, numbered and labelled with the patch around it."
question: arrangement
about: forest
scale: patch
declares: []
uses: [patches, holes]
result:
  - {object: holes, layer: neighbour, text: "on land enclosed by patch {v}"}
---

Holes are the land that a patch encloses: glades, windthrows, clearings, the space inside
a loop. Filling the patches and subtracting the original leaves the holes;
each hole is then numbered, and labelled with the patch around it.

The forest patches are numbered first. The three large ones get the numbers
3 (north-east), 4 (north-west) and 5 (south-west):

```{r}
library(mosaik)
patches <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch")
```

```{r}
holes <- patches |>
  mdf_fill(layer = "forest", add = "filled") |>
  mdf_blend(layers = c("filled", "forest"), fun = "-", add = "hole") |>
  mdf_componentise(layer = "hole", add = "hole_id") |>
  mdf_summarise(by = "hole_id", fun = "unique", neighbours = TRUE,
                layer = "patch", add = "neighbour")
```

**Check.** The map has three holes in two patches: an old windthrow in patch
4, and in patch 3 a square glade and the land inside its loop.

```{r, fig.alt = "The numbered holes, and the number of the patch around each"}
msk_vis(holes, .layer("hole_id"), .layer("neighbour"))
```

```{r, include = FALSE}
ids <- msk_pull(holes, "hole_id")
around <- tapply(msk_pull(holes, "neighbour"), ids, unique)
stopifnot(length(around) == 3, sort(around) == c(3, 3, 4))
```
