---
title: "Indentations"
summary: "Bays cut into a patch from its edge, the shape a clear-cut leaves, found by closing the patch."
question: arrangement
about: forest
scale: patch
declares: ["the widest indentation to count (k)"]
uses: [patches, holes]
result:
  - {object: cut, layer: notch, text: "in an indentation of patch 4"}
---

An indentation is a bay cut into a patch from its edge. It is the shape a
clear-cut typically leaves: cut in from the edge, so it is not enclosed and is
no hole. The shape alone does not tell what made it, whatever the land in it
is now, but it marks the likely clear-cuts of a forest from the forest's
outline alone. Closing the patch finds it: growing the patch by `k` cells and
shrinking it again by `k` cells fills every indentation narrower than about
`2k` cells and leaves the rest of the outline where it was. Growing by `k`
cells is one dilation with a square `2k + 1` cells wide. What closing adds,
apart from the holes, are the indentations.

The recipe works on one patch at a time, so that neighbouring patches do not
grow into each other. Here patch 4, the large patch in the north-west:

```{r}
library(mosaik)
one <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_filter(patch == 4, add = "one") |>
  mdf_fill(layer = "one", add = "one_filled")

k <- 3
sq <- msk_struct("square", width = 2 * k + 1, height = 2 * k + 1)
cut <- one |>
  mdf_dilate(struct = sq, layer = "one", add = "closed") |>
  mdf_erode(struct = sq, layer = "closed") |>
  mdf_filter(closed == 1 & one_filled == 0, add = "notch")
```

```{r, fig.alt = "Patch 4 and what closing it by three cells adds"}
msk_vis(cut, .layer("one"), .layer("notch"))
```

**Check.** The check repeats the recipe for `k` from 1 to 4 and counts, for
each `k`, how many cells closing adds to the patch outside its holes. The
clear-cut in patch 4 is six cells wide, so it only closes from `k = 3` on,
since `2k` must reach six. With `k = 1` and `k = 2`, the count should stay
small, made up of the little bays along the outline. At `k = 3` it should jump
by the size of the clear-cut, about 30 cells, and at `k = 4` grow only a
little more:

```{r}
added <- sapply(1:4, function(k) {
  sq <- msk_struct("square", width = 2 * k + 1, height = 2 * k + 1)
  closed <- one |>
    mdf_dilate(struct = sq, layer = "one") |>
    mdf_erode(struct = sq, layer = "one")
  sum(msk_pull(closed, "one") == 1 & msk_pull(one, "one_filled") == 0)
})
added
```

```{r, include = FALSE}
stopifnot(added[3] - added[2] > 20, added[4] - added[3] < added[3] - added[2])
```
