---
title: "Interrupted corridors"
summary: "Patches separated only by a narrow break, such as a road or a river."
question: arrangement
about: forest
scale: patch
declares: ["the widest break to bridge (cells)"]
uses: [patches]
result:
  - {object: joined, layer: joined, text: "its patch is one of {v} patches separated only by narrow breaks", omit: 1}
---

A road or a river can cut a corridor, but the break is narrow. Closing the
forest by one cell bridges such breaks; patches that were apart and are joined
afterwards were separated only by a narrow break.

```{r}
library(mosaik)
joined <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_dilate(layer = "forest", add = "closed") |>
  mdf_erode(layer = "closed") |>
  mdf_componentise(connectivity = 8L, layer = "closed", add = "group") |>
  mdf_summarise(by = "group", fun = "n_distinct", layer = "patch",
                add = "joined") |>
  mdf_filter(forest == 1, value = TRUE, layer = "joined")
```

The large patches are drawn in colour: 3 in the north-east, 4 in the
north-west and 5 in the south-west.

```{r, fig.width = 9, fig.alt = "Forest patches, the groups they form after closing, and how many patches each group joins"}
cols <- rep("grey75", 17)
cols[3:5] <- c("#e66101", "#2166ac", "#1b7837")
msk_vis(joined, .layer("patch", colours = cols), .layer("group"),
        .layer("joined"))
```

**Check.** The river cuts the corridor between patches 4 and 3, and the road
separates patch 4 from patch 5. Across these breaks, the three patches become
one group. Three smaller patches lie within one cell of patch 5 and join the
same group: the strip of forest south of it and two fragments to its west. The
group should therefore hold six patches:

```{r}
g <- unique(msk_pull(joined, "group")[msk_pull(joined, "patch") %in% 3:5])
members <- sort(unique(msk_pull(joined, "patch")[msk_pull(joined, "group") %in% g]))
length(g)
members
```

```{r, include = FALSE}
stopifnot(length(g) == 1, length(members) == 6, all(3:5 %in% members))
```
