---
title: "Islets"
summary: "Patches of forest too small or too narrow to hold any core."
question: arrangement
about: forest
scale: patch
declares: ["the edge width (cells)"]
uses: [patches, core]
result:
  - {object: islets, layer: islet, text: "on an islet: a patch too small or too narrow to hold any core"}
---

A patch without any core is edge throughout: whatever lies outside reaches
every part of it. Such patches can still carry species that move through the
land, as stepping stones, but they hold no interior habitat of their own. The
recipe asks of every patch whether any of its cells is core, with the core
two or more cells from the edge:

```{r}
library(mosaik)
islets <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_filter(depth >= 2, add = "core") |>
  mdf_summarise(by = "patch", fun = "any", layer = "core", add = "has_core") |>
  mdf_filter(forest == 1 & has_core == 0, add = "islet") |>
  mdf_summarise(by = "patch", fun = "n", layer = "patch", add = "size")
```

```{r, fig.alt = "The forest, its core, and the patches without core"}
msk_vis(islets, .layer("forest"), .layer("core"), .layer("islet"))
```

**Check.** Being an islet is a matter of shape as much as of size. The
smallest patches are islets, and so is the strip in the south, 30 cells long
but too narrow for any core. The patch at the south-eastern corner is smaller,
26 cells, but compact, and holds core:

```{r}
p <- msk_pull(islets, "patch")
patches <- data.frame(size = tapply(msk_pull(islets, "size"), p, unique),
                      islet = tapply(msk_pull(islets, "islet"), p, max))
patches[order(patches$size, decreasing = TRUE), ][1:6, ]
```

```{r, include = FALSE}
stopifnot(all(patches$islet[patches$size < 12] == 1),
          patches["15", "islet"] == 1, patches["17", "islet"] == 0,
          sum(patches$islet) == 13)
```
