---
title: "Linear elements"
summary: "Hedgerows and corridors told apart from compact blocks of forest."
question: arrangement
about: forest
scale: patch
declares: []
uses: [patches, corridors]
result:
  - {object: linear, layer: shape, text: "{label}", labels: {"1": "its patch is a compact block", "2": "its patch is elongated", "3": "its patch is a linear element, such as a hedgerow"}}
---

Hedgerows and corridors are long and narrow, blocks of forest are compact.
The share of a patch that lies on its skeleton separates them: in a strip one
or two cells wide almost every cell is skeleton, in a compact patch few are.

```{r}
library(mosaik)
linear <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_componentise(connectivity = 8L, layer = "forest", add = "patch") |>
  mdf_skeletonise(layer = "forest", add = "skeleton") |>
  mdf_summarise(by = "patch", fun = "sum", layer = "skeleton", add = "spine") |>
  mdf_summarise(by = "patch", fun = "n", layer = "patch", add = "size") |>
  mdf_blend(layers = c("spine", "size"), fun = "/", add = "linearity") |>
  mdf_categorise(breaks = c(1/3, 0.7), layer = "linearity", add = "shape")
```

`shape` sorts the patches by their linearity: below a third they are compact
blocks (1), above 0.7 linear elements such as hedgerows (3), and elongated in
between (2).

```{r, fig.alt = "Forest patches, coloured by the share of their cells on the skeleton"}
msk_vis(linear, .layer("linearity"), .layer("shape"))
```

**Check.** The large patches 3, 4 and 5 are compact, so few of their cells
lie on the skeleton: they should score below a third. The hedgerows of the
bocage are a few cells long and one or two wide, so most of their cells do:
they should score above 0.7. Single cells are trivially all skeleton, so only
patches of at least four cells are compared:

```{r}
patch <- msk_pull(linear, "patch")
size <- tapply(msk_pull(linear, "size"), patch, unique)
linearity <- tapply(msk_pull(linear, "linearity"), patch, unique)
round(sort(linearity[size >= 4]), 2)
```

```{r, include = FALSE}
stopifnot(all(linearity[c("3", "4", "5")] < 1/3),
          sum(linearity[size >= 4] > 0.7) >= 4,
          all(size[size >= 4 & linearity > 0.7] <= 11))
```
