---
title: "Parallel structures"
summary: "Where one linear structure keeps a constant distance to another, like a road beside a river."
question: arrangement
about: road
scale: cell
declares: ["the distance that counts as beside (cells)"]
uses: [class, distance]
result:
  - {object: parallel, layer: beside, text: "on the road where it runs beside the river"}
---

Two linear structures run in parallel where one keeps a constant distance to
the other, like a road beside a river. The distance from the river, read along
the road, shows where it does.

```{r}
library(mosaik)
parallel <- landscape |>
  mdf_binarise(match = 1, layer = "cover", add = "river") |>
  mdf_distance(source = "foreground", layer = "river", add = "to_river") |>
  mdf_binarise(match = 35, layer = "cover", add = "road") |>
  mdf_filter(road == 1 & to_river <= 3, value = TRUE, add = "beside")
```

`beside` is the stretch of road within three cells of the river. It is drawn
over the river, so the two can be compared; cells of the road elsewhere are
empty (`NA`) and let the river show through:

```{r, fig.alt = "The road, and the river with the part of the road that runs beside it"}
msk_vis(parallel, .layer("road"),
        .layer("river", panel = "river and road beside it",
               colours = c("grey92", "#2b6cb0")),
        .layer("beside", panel = "river and road beside it",
               colours = "#e66101", legend = FALSE))
```

**Check.** The road follows the river from the southern edge, first with one
cell between them (a distance of 2), then with two (a distance of 3). Where
the river bends, the road steps sideways, and the cells at each step and next
to each bend come closer than that (1 to 2.8). The distances along the road
show all of these:

```{r}
d <- msk_pull(parallel, "to_river")[msk_pull(parallel, "beside") %in% 1]
table(distance = round(d, 1))
```

```{r, include = FALSE}
tab <- table(round(d, 1))
stopifnot(all(d >= 1 & d <= 3),
          all(names(sort(tab, decreasing = TRUE))[1:2] %in% c("2", "3")))
```
