---
title: "The structure of a protected area"
summary: "How much of a declared area is forest, how much of that is core, and into how many patches it falls."
question: arrangement
about: area
scale: zone
declares: ["the area", "the edge width (cells)"]
uses: [protected-area, patches, core]
result:
  - {object: protected, layer: area, text: "inside the protected area"}
  - {object: protected, layer: core_share, text: "{v} % of the protected area is forest core", percent: true}
---

A protected area is usually given as a polygon. `msk_rasterise()` writes it
onto the grid of the map, and from there its structure is measured like
anything else: how much of it is forest, how much of that forest is core,
more than two cells from any edge, and into how many patches the forest falls.

```{r}
library(mosaik)
area <- data.frame(x = c(3, 25, 25, 3), y = c(25, 25, 53, 53), id = 1)

protected <- landscape |>
  msk_rasterise(geom = area, name = "area") |>
  mdf_replace(old = NA, new = 0, layer = "area") |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_filter(area == 1 & forest == 1, add = "area_forest") |>
  mdf_erode(struct = msk_struct("square", width = 5), layer = "area_forest",
            add = "area_core") |>
  mdf_componentise(connectivity = 8L, layer = "area_forest",
                   add = "area_patch") |>
  mdf_summarise(by = "area", fun = "mean", layer = "forest",
                add = "forest_share") |>
  mdf_summarise(by = "area", fun = "mean", layer = "area_core",
                add = "core_share") |>
  mdf_summarise(by = "area", fun = "n_distinct", layer = "area_patch",
                add = "patches")
```

```{r, fig.alt = "The protected area, its forest, and the core of that forest"}
msk_vis(protected, .layer("area"), .layer("area_forest"), .layer("area_core"))
```

Each summary is written onto every cell of the area, so it is read from any of
them:

```{r}
in_area <- msk_pull(protected, "area") == 1
structure <- c(forest = unique(msk_pull(protected, "forest_share")[in_area]),
               core = unique(msk_pull(protected, "core_share")[in_area]),
               patches = unique(msk_pull(protected, "patches")[in_area]))
structure
```

**Check.** The area covers patch 4, but the patch is not all forest: a windthrow,
the fresh clear-cut and the road lie inside it, and the rectangle reaches into
open land around it. The forest should therefore cover less than half of the
area, the core much less. The road cuts the patch, so the forest falls into
more than one patch.

```{r, include = FALSE}
stopifnot(structure[["forest"]] < 0.5,
          structure[["core"]] < structure[["forest"]] / 2,
          structure[["patches"]] > 1)
```

The recipe "What the road costs" measures how much of this structure is due
to the road.
