---
title: "Riparian strips"
summary: "The share of natural land in a strip along each stretch of a river."
question: pressure
about: river
scale: zone
declares: ["which land cover counts as natural", "the width of the strip (cells)", "the stretches"]
uses: [class, distance, zones]
result:
  - {object: riparian, layer: natural_share, text: "{v} % of the strip along this stretch of river is natural land", percent: true}
---

Land use right next to a river acts on it: fertiliser and soil wash in, and
shade and shelter go missing. Rules for farming near water often require a
strip along rivers that is kept free of intensive use. The recipe finds the
strip, splits the river into stretches, and computes the share of natural
land in the strip along each stretch.

Which land cover counts as natural is a declaration. Here it is the river,
extensive grassland, fallow and shrub, wetland, and forest:

```{r}
library(mosaik)
natural <- c(1, 24, 27, 41, 47)

riparian <- landscape |>
  mdf_binarise(match = 1, layer = "cover", add = "river") |>
  mdf_distance(layer = "river", add = "to_river") |>
  mdf_filter(to_river > 0 & to_river <= 3, add = "strip") |>
  mdf_binarise(match = natural, layer = "cover", add = "natural") |>
  drw_gradient(type = "planar", angle = 90, name = "north") |>
  mdf_categorise(n = 4, layer = "north", add = "stretch") |>
  mdf_blend(layers = c("strip", "stretch"), fun = "*", add = "zone") |>
  mdf_summarise(by = "zone", fun = "mean", layer = "natural",
                add = "natural_share")
```

The strip is three cells wide on either side. The stretches are four bands of
equal height from south to north, made by cutting a gradient that rises
northwards into four classes. Multiplying the strip (0 or 1) by the stretch
number gives each cell of the strip the number of its stretch, and 0 to every
other cell, which `mdf_summarise()` leaves out:

```{r, fig.alt = "The strip along the river, its four stretches, and the share of natural land along each"}
msk_vis(riparian, .layer("strip"), .layer("zone"), .layer("natural_share"))
```

**Check.** In the southern third of the map, a road runs beside the river,
first one and then two cells away from it, so the strip along the southern
stretch is largely road and its natural share should be the lowest. In the
north, the river runs along a large intensive meadow, so the northern stretch
should be low as well. The two stretches in between run through land of mixed
use:

```{r}
zone <- msk_pull(riparian, "zone")
shares <- tapply(msk_pull(riparian, "natural_share")[zone > 0], zone[zone > 0],
                 unique)
round(shares, 2)
```

```{r, include = FALSE}
stopifnot(which.min(shares) == 1, shares[["4"]] < min(shares[c("2", "3")]))
```

Whether a stretch complies is a further declaration: a rule might demand that
all of the strip is natural, or a share of it.
