---
title: "What each stretch of road costs"
summary: "Where along the road the forest core is lost, stretch by stretch."
question: cause
about: road
scale: zone
declares: ["the length of a stretch", "the edge width (cells)", "the counterfactual: forest in place of the stretch"]
uses: [declared, core, zones]
result:
  - {object: stretches, layer: cost, text: "this stretch of road costs {v} cells of forest core", omit: 0}
---

The recipe on what the road costs treats the road as one thing. Where along
the road the cost arises is a different question, and the one a decision
about the road needs: which stretch to move, to bridge, or to leave alone.
The road is cut into stretches, and each stretch in turn is replaced by forest
while the rest of the road stays. The core gained is what that stretch costs.

The stretches are the parts of the road within squares of six by six cells:

```{r}
library(mosaik)
stretches <- landscape |>
  mdf_binarise(match = 47, layer = "cover", add = "forest") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_binarise(thresh = 2, layer = "depth", add = "core") |>
  drw_pattern(type = "checkerboard", frequency = 6, n = 0, name = "square") |>
  mdf_binarise(match = 35, layer = "cover", add = "road") |>
  mdf_blend(layers = c("road", "square"), fun = "*", add = "stretch")

# the core habitat of a forest layer "without", recorded once as a recipe
core_of <- mdf_distance(source = "background", layer = "without", add = "d") |>
  mdf_binarise(thresh = 2, layer = "d", add = "c")

ids <- sort(setdiff(unique(msk_pull(stretches, "stretch")), 0))
core_now <- sum(msk_pull(stretches, "core"))

# for each stretch, the core habitat with forest in place of that stretch only
gained <- vapply(ids, function(k) {
  counterfactual <- stretches |>
    mdf_filter(forest == 1 | stretch == k, add = "without") |>
    mdf(core_of)
  sum(msk_pull(counterfactual, "c")) - core_now
}, numeric(1))

stretches <- stretches |>
  mdf_replace(old = ids, new = gained, layer = "stretch", add = "cost") |>
  mdf_filter(road == 1, value = TRUE, layer = "cost")
```

```{r, fig.alt = "The road, and the forest core each stretch of it costs"}
msk_vis(stretches, .layer("road"), .layer("cost"))
```

**Check.** Most of the road runs through open land and beside the river,
where replacing it by forest gains no core. Only the stretches through patch 4
should cost any:

```{r}
data.frame(stretch = ids, core_cost = gained)[gained > 0, ]
```

Adding up the stretches does not give the cost of the whole road. With forest
in place of all of it at once, the core gained is:

```{r}
whole <- stretches |>
  mdf_filter(forest == 1 | road == 1, add = "without") |>
  mdf(core_of)
c(stretches_added_up = sum(gained),
  whole_road = sum(msk_pull(whole, "c")) - core_now)
```

```{r, include = FALSE}
stopifnot(sum(gained > 0) <= 4, sum(gained) > 0,
          sum(msk_pull(whole, "c")) - core_now > sum(gained))
```

The whole road costs more than its stretches added up. Losses of core do not
add up: two cuts close together destroy core that neither would destroy
alone, so removing one stretch while its neighbours remain gives back less
than its share.
