---
title: "What the road costs"
summary: "The core habitat a road destroys in a protected area: the undisturbed forest interior lost along both sides of it."
question: cause
about: road
scale: zone
declares: ["the area", "the edge width (cells)", "the counterfactual: forest in place of the road where it cuts through the forest"]
uses: [protected-area, declared, patches, core]
result:
  - {object: road, layer: lost, text: "core habitat the road destroys: undisturbed forest interior without the road"}
---

Many forest species need core habitat: forest far enough from any edge to be
free of the light, wind, noise and predators that come in from outside. Here
that is forest two or more cells from an edge. A road through a forest
takes far more core habitat than its own surface, because it makes a new
edge on both sides: along the road, a band two cells deep on either side
stops being core. A road one or two cells wide therefore destroys core in a
band about six cells wide.

How much of the core habitat of a protected area the road destroys needs a
counterfactual: the same area without the road. Here it is declared as forest
in place of the road, but only where the road cuts through the forest. Where
the road runs along the edge of the forest or through open land, forest in
its place would widen the forest beyond what it ever was. The cells where the
road cuts through are those that closing the forest by one cell covers:

```{r}
library(mosaik)
area <- data.frame(x = c(3, 25, 25, 3), y = c(25, 25, 53, 53), id = 1)
square <- msk_struct("square", width = 3)
wide <- msk_struct("square", width = 5)

road <- landscape |>
  msk_rasterise(geom = area, name = "area") |>
  mdf_replace(old = NA, new = 0, layer = "area") |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_dilate(struct = square, layer = "forest", add = "closed") |>
  mdf_erode(struct = square, layer = "closed") |>
  mdf_filter(area == 1 & cover == 35 & closed == 1, add = "through") |>
  mdf_filter(area == 1 & forest == 1, add = "with_road") |>
  mdf_filter(area == 1 & (forest == 1 | through == 1), add = "without_road")
```

The core habitat is then measured on both, and `lost` is the core habitat
that exists only without the road: what the road destroys. It includes the
road's own cells, which would be core habitat had the forest not been cut.

```{r}
road <- road |>
  mdf_erode(struct = wide, layer = "with_road", add = "core") |>
  mdf_erode(struct = wide, layer = "without_road", add = "core_without") |>
  mdf_componentise(connectivity = 8L, layer = "with_road", add = "patch") |>
  mdf_componentise(connectivity = 8L, layer = "without_road",
                   add = "patch_without") |>
  mdf_filter(core_without == 1 & core == 0, add = "lost")
```

```{r, fig.alt = "The road where it cuts through the forest, the core habitat as it would be without the road, and the core habitat the road destroys"}
msk_vis(road, .layer("through"), .layer("core_without"), .layer("lost"))
```

**Check.** Where the road crosses the forest, the core habitat on both sides
of it is lost, so without the road there should be more core habitat, and
the two parts of the patch the road separates should be one again:

```{r}
cost <- c(core = sum(msk_pull(road, "core")),
          core_without_road = sum(msk_pull(road, "core_without")),
          patches = length(unique(na.omit(msk_pull(road, "patch")))),
          patches_without_road =
            length(unique(na.omit(msk_pull(road, "patch_without")))))
cost
```

```{r, include = FALSE}
stopifnot(cost[["core_without_road"]] > cost[["core"]],
          cost[["patches_without_road"]] == cost[["patches"]] - 1)
```

Where the road cuts through the forest of the area, it covers only a few
cells, but the core habitat it destroys is nearly twice that, since most of
it lies on forest that still stands beside the road:

```{r}
c(road_through_forest = sum(msk_pull(road, "through")),
  core_habitat_lost = sum(msk_pull(road, "lost")))
```

```{r, include = FALSE}
stopifnot(sum(msk_pull(road, "lost")) > 1.5 * sum(msk_pull(road, "through")))
```

The difference holds only against the declared counterfactual: had the land
under the road been something other than forest, the difference would be
another. The recipe "What each stretch of road costs" shows where along the
road the core habitat is lost.
