---
title: "Sharp and gradual edges"
summary: "Forest edges with a mantle of lower trees told apart from those where tall trees stand right at the edge."
question: state
about: forest
scale: cell
declares: ["the drop in height that makes an edge gradual (m)"]
uses: [canopy, ecotone, distance]
result:
  - {object: sharpness, layer: sharp, text: "on a sharp forest edge"}
  - {object: sharpness, layer: gradual, text: "on a gradual forest edge"}
---

An old forest edge has grown a mantle: shrubs and lower trees, so the canopy
rises gradually into the forest. A fresh edge, such as a clear-cut or a road
recently built through the forest, is sharp: the trees that stood inside the
forest stand at the edge at their full height. The canopy height of the
outermost forest cells, compared with that of the interior, tells the two
apart. An edge cell counts as gradual when its canopy lies 6 m or more below
the median of the interior.

```{r}
library(mosaik)
edge <- landscape |>
  mdf_filter(cover == 47, add = "forest") |>
  mdf_distance(source = "background", layer = "forest", add = "depth") |>
  mdf_filter(depth >= 1 & depth < 1.5, add = "rim")
level <- median(msk_pull(edge, "canopy")[msk_pull(edge, "depth") > 3])
sharpness <- edge |>
  mdf_filter(rim == 1 & canopy > level - 6, add = "sharp") |>
  mdf_filter(rim == 1 & canopy <= level - 6, add = "gradual")
```

```{r, fig.width = 9, fig.alt = "Canopy height, and the forest edge cells where the edge is sharp and where it is gradual"}
msk_vis(sharpness, .layer("canopy"), .layer("sharp"), .layer("gradual"))
```

**Check.** The sharp edges are those along the road and the fresh clear-cuts.
Most sharp edge cells therefore lie next to the road or to fallow and shrub,
while few of the gradual ones do:

```{r}
fresh <- landscape |>
  mdf_filter(cover %in% c(35, 27), add = "fresh") |>
  mdf_dilate(struct = msk_struct("square", width = 3, height = 3),
             layer = "fresh")
near <- msk_pull(fresh, "fresh") == 1
sharp <- msk_pull(sharpness, "sharp") == 1
gradual <- msk_pull(sharpness, "gradual") == 1
c(sharp = mean(near[sharp]), gradual = mean(near[gradual]))
```

```{r, include = FALSE}
stopifnot(mean(near[sharp]) > 0.6, mean(near[gradual]) < 0.25)
```
